---
title: "P1_Data_Exploration"
output: html_document
date: "2026-02-08"
editor_options: 
  chunk_output_type: console
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
library(dplyr)
library(naniar)
library(table1)
dat_raw <- read.csv(here::here("P1/DataRaw", "hiv_6624_final.csv"))
```

# Glimpsing Data

Initial overview:
```{r}
summary(dat_raw)
glimpse(dat_raw)
```

Missing data:
```{r}
naniar::gg_miss_var(dat_raw)
naniar::vis_miss(dat_raw)
```
Is there a reason for this missing data? There is some for the VLOAD and LEU3N, as well as a minimal amount for hard drug use, which seem to be the focus for the analysis.

Table 1:
```{r}
table1(~ hard_drugs + hivpos + ART + everART + VLOAD + LEU3N + AGG_PHYS + AGG_MENT + age | years, dat_raw)
```
Dropoff year 0 to year 2? 

```{r}
hist(dat_raw$VLOAD); hist(dat_raw$LEU3N); hist(dat_raw$AGG_PHYS); hist(dat_raw$AGG_MENT)
hist(log10(dat_raw$VLOAD)); hist(sqrt(dat_raw$LEU3N)); hist(log(101-dat_raw$AGG_PHYS)); hist(log(101-dat_raw$AGG_MENT))
```
What are reasonable values/ranges of these values? What extremes are unrealistic? These align with the codebook allowable values


# Potential analysis approach:

Maybe use adjusted mean difference in year 2 outcome between baseline hard drug users vs. non users, controlling for baseline values, baseline covariates (age, BMI, smoking, education, race)

Frequentist: linear regression models, one for each outcome. Model year 2 outcomes predicted by hard drug use, covariates, and baseline value
- VLOAD (logged)
- LEU3N (maybe sqrt transformation?)
- AGG_PHYS
- AGG_MENT

Treat adherence as a mediator, comparing predictor from model without and with adherence (reduction in slope suggests mediation)

Potentially use FDR control for p-values among separate models 

Compare baseline characteristics and hard drug use between those with and without year 2 data to try and identify possible bias (regression assuming MAR)

Bayesian: similar to frequentist (linear regression) but with weekly informative/vague priors
- normal priors on regression coefficients
- half normal prior on residual SD
- stan program with design matrix x and outcome y
- posterior sampling with cmdstanr
- diagnostics, posterior summaries, and WAIC/LOO

Clarifying questions:
- What strategies for collapsing race and education categories would you recommend? 
  e.g. end with white (non-hispanic), black (non-hispanic), other (742, 345, 134)
      or, white (non-hispanic), black (non-hispanic), hispanic, other (742, 345, 121, 74) - less ideal
  e.g. end with high school and below, some college, college and higher (291, 434, 496)
- Do you want a formal mediation analysis or just a model comparison for adherence? 


# Investigation into Relevant Data

```{r}
dat_sub <- dat_raw %>%
  filter(years %in% c(0, 2)) %>%
    # Filter to baseline and year 2
  mutate(
    race_group = case_when(
      RACE == 1 ~ "White_nonHisp",
      RACE == 3 ~ "Black_nonHisp",
      RACE %in% c(2, 4, 5, 6, 7, 8) ~ "Other",
      TRUE ~ NA_character_
    )
  ) %>%
  mutate(
    educ_group = case_when(
      EDUCBAS %in% c(1, 2, 3) ~ "HS_and_below",
      EDUCBAS == 4 ~ "Some_College",
      EDUCBAS %in% c(5, 6, 7) ~ "College_and_above",
      TRUE ~ NA_character_
    )
  )

baseline_drugs <- dat_sub %>%
  filter(years == 0) %>%
  select(newid, hard_drugs)
    # Identify baseline drug use by id
dat_sub <- dat_sub %>%
  left_join(baseline_drugs, by = "newid", suffix = c("", "_baseline"))
    # Add baseline drug use to dataset

table1(~ VLOAD + LEU3N + AGG_PHYS + AGG_MENT + age + BMI + SMOKE + educ_group + race_group | hard_drugs, dat_sub %>% filter(years == 0))
table1(~ VLOAD + LEU3N + ADH| hard_drugs_baseline, dat_sub %>% filter(years == 2))
```

```{r}
dat_2 <- dat_sub %>%
  filter(years == 0)
par(mfrow = c(2,4))
hist(dat_2$VLOAD, main = "VLOAD Year 2", xlab = "VLOAD"); hist(dat_2$LEU3N, main = "LEU3N Year 2", xlab = "LEU3N"); hist(dat_2$AGG_PHYS, main = "Physical Year 2", xlab = "Physical QoL"); hist(dat_raw$AGG_MENT, main = "Mental Year 2", xlab = "Mental QoL")
hist(log10(dat_2$VLOAD), main = "log10(VLOAD) Year 2", xlab = "log10(VLOAD)"); hist(sqrt(dat_2$LEU3N), main = "sqrt(LEU3N) Year 2", xlab = "sqrt(LEU3N)"); hist(log(101-dat_2$AGG_PHYS), main = "log(101-Physical) Year 2", xlab = "log(101-Physical)"); hist(log(101-dat_2$AGG_MENT), main = "log(101-Mental) Year 2", xlab = "log(101-Mental)")
```


