---
title: "P1_Draft_Analysis"
output: html_document
date: "2026-02-25"
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
library(dplyr)
library(tidyr)
library(tidyverse)
library(purrr)
library(ggplot2)
library(table1)
library(cmdstanr)
library(posterior)
library(tidybayes)
library(tibble)
library(mcmcse)
library(gt)
dat_raw <- read.csv(here::here("P1/DataRaw", "hiv_6624_final.csv"))
```

```{r}
dat_sub <- dat_raw %>%
  filter(
    years %in% c(0, 2)
  ) %>%
    # Filter to baseline and year 2
  mutate(
    race_group = case_when(
      RACE == 1 ~ "White nonHisp",
      RACE %in% 2:8 ~ "Other",
      TRUE ~ NA_character_
    ), # Collapse race
    educ_group = case_when(
      EDUCBAS %in% 1:4 ~ "No College Degree",
      EDUCBAS %in% 5:7 ~ "College Degree and above",
      TRUE ~ NA_character_
    ), # Collapse education
    smoke_group = case_when(
      SMOKE %in% c(1, 2) ~ "Non or Former Smoker",
      SMOKE == 3 ~ "Current Smoker",
      TRUE ~ NA_character_
    ), # Collapse smoking status
    adh_group = case_when(
      ADH %in% c(1, 2) ~ "95% and Above",
      ADH %in% c(3, 4) ~ "<95%",
      TRUE ~ NA_character_
    ), # Collapse adherence 
    hard_drugs = case_when(
      hard_drugs == 0 ~ "No",
      hard_drugs == 1 ~ "Yes",
      TRUE ~ NA_character_
    ),
      # Add labels to hard drugs
    BMI = if_else(
      BMI < 0 | BMI > 250, NA_real_, BMI
    ), # Exclude implausible BMI values
    log10VLOAD = log10(VLOAD)
      # Convert VLOAD to log10 scale
  )

dat_wide <- dat_sub %>%
  pivot_wider(
    id_cols = newid,
    names_from = years,
    values_from = c(
      log10VLOAD, LEU3N, AGG_PHYS, AGG_MENT,
      age, BMI, smoke_group, educ_group, race_group,
      hard_drugs, adh_group
    ),
    names_glue = "{.value}_{years}"
  )

required_outcomes <- c("log10VLOAD_0", "log10VLOAD_2", "LEU3N_0", "LEU3N_2",
                       "AGG_PHYS_0", "AGG_PHYS_2", "AGG_MENT_0", "AGG_MENT_2")
required_predictors <- c("age_0", "BMI_0", "smoke_group_0", "educ_group_0",
                         "race_group_0", "hard_drugs_0", "adh_group_2")

dat_wide <- dat_wide %>%
  filter(
    complete.cases(across(c(required_outcomes, required_predictors)))
  ) %>% # Filter to those without missing outcomes or predictors
  select(
    newid, all_of(required_outcomes), all_of(required_predictors)
  ) # Only select relevant columns
  # 463 final sample size

table1a <- table1(~ log10VLOAD_0 + LEU3N_0 + AGG_PHYS_0 + AGG_MENT_0 + age_0 + BMI_0 + smoke_group_0 + educ_group_0 + race_group_0 | hard_drugs_0, dat_wide, caption="Baseline Characteristics by Baseline Hard Drug Use")
table1b <- table1(~ log10VLOAD_2 + LEU3N_2 + AGG_PHYS_2 + AGG_MENT_2 + adh_group_2 | hard_drugs_0, dat_wide, caption="Year 2 Characteristics by Baseline Hard Drug Use")
```

# TODO: analysis steps

- collapse ADH to upper 2 and lower 2 groups (>95% and <95%)
- reduce to 463 individuals with complete covariate data
- use dplyr to transform dat_sub so that you have y2 log10VLOAD/other outcomes and baseline hard drug use/covariates/y2 adherence all available in one single row
- model with simple linear regression with separate models for all 4 outcomes
  - store coefficients, unadjusted pvals, 95% CIs
  - model with adherence and store % change in outcome (are there differences in adherence/relationship?)
- get STAN to run and model with bayesian linear regression with N(0,1000) for regression coefficients and HN(0,1000) for error. 
  - store posterior mean, HDPI, posterior probability (use clinically meaningful differences)
  - model with adherence and store % change in outcome
- Make table of outcomes that combines frequentist and bayesian outcomes previously stored (table 3)
- Consider what figures or any additional tables that you want to include
  - perhaps separate table of adherence % changes to table 3
------------------------------------------------------------------------------
  - perhaps some figure to visualize significant outcomes by hard drug status/somehow visualize the models (think boxplot of each outcome with 2 for each color coded by hard drug use, each outcome as a panel)
- write report 
- make presentation


# Frequentist Analysis 

```{r}
mod_vload <- lm(log10VLOAD_2 ~ hard_drugs_0 + log10VLOAD_0 + age_0 + BMI_0 +
                  smoke_group_0 + educ_group_0 + race_group_0, data = dat_wide)
mod_cd4 <- lm(LEU3N_2 ~ hard_drugs_0 + LEU3N_0 + age_0 + BMI_0 + smoke_group_0 +
                educ_group_0 + race_group_0, data = dat_wide)
mod_phys <- lm(AGG_PHYS_2 ~ hard_drugs_0 + AGG_PHYS_0 + age_0 + BMI_0 +
                 smoke_group_0 + educ_group_0 + race_group_0, data = dat_wide)
mod_ment <- lm(AGG_MENT_2 ~ hard_drugs_0 + AGG_MENT_0 + age_0 + BMI_0 +
                 smoke_group_0 + educ_group_0 + race_group_0, data = dat_wide)
```

```{r}
mod_vload_adh <- lm(log10VLOAD_2 ~ hard_drugs_0 + log10VLOAD_0 + age_0 + BMI_0 +
                      smoke_group_0 + educ_group_0 + race_group_0 + adh_group_2,
                    data = dat_wide)
mod_cd4_adh <- lm(LEU3N_2 ~ hard_drugs_0 + LEU3N_0 + age_0 + BMI_0 + 
                    smoke_group_0 + educ_group_0 + race_group_0 + adh_group_2,
                  data = dat_wide)
mod_phys_adh <- lm(AGG_PHYS_2 ~ hard_drugs_0 + AGG_PHYS_0 + age_0 + BMI_0 +
                     smoke_group_0 + educ_group_0 + race_group_0 + adh_group_2,
                   data = dat_wide)
mod_ment_adh <- lm(AGG_MENT_2 ~ hard_drugs_0 + AGG_MENT_0 + age_0 + BMI_0 +
                     smoke_group_0 + educ_group_0 + race_group_0 + adh_group_2,
                   data = dat_wide)
```

```{r}
summary(mod_vload)
summary(mod_vload_adh)
```
```{r}
summary(mod_cd4)
summary(mod_cd4_adh)
```
```{r}
summary(mod_phys)
summary(mod_phys_adh)
```
```{r}
summary(mod_ment)
summary(mod_ment_adh)
```
Summary table
```{r}
summarise_lm_beta2 <- function(mod) {
  
  sm <- summary(mod)
  coefs <- sm$coefficients
  
  est <- coefs["hard_drugs_0Yes", "Estimate"]
  se  <- coefs["hard_drugs_0Yes", "Std. Error"]
  p   <- coefs["hard_drugs_0Yes", "Pr(>|t|)"]
  
  ci_low  <- est - 1.96 * se
  ci_high <- est + 1.96 * se
  
  tibble(
    Parameter = "hard_drugs_0Yes",
    PE = est,
    CI_2.5 = ci_low,
    CI_97.5 = ci_high,
    p_value = p
  )
}
mod_list <- list(
  `Viral Load (log10 HIV copies/mL blood)` = mod_vload,
  `Viral Load (log10 HIV copies/mL blood)_adh` = mod_vload_adh,
  `CD4+ T Cell Count` = mod_cd4,
  `CD4+ T Cell Count_adh` = mod_cd4_adh,
  `Physical Quality of Life (SF-36 Aggregate)` = mod_phys,
  `Physical Quality of Life (SF-36 Aggregate)_adh` = mod_phys_adh,
  `Mental Quality of Life (SF-36 Aggregate)` = mod_ment,
  `Mental Quality of Life (SF-36 Aggregate)_adh` = mod_ment_adh
)
tab_freq <- purrr::map_df(mod_list, summarise_lm_beta2, .id = "Model") %>%
  mutate(
    CI = sprintf("(%0.3f, %0.3f)", CI_2.5, CI_97.5)
      # Combine CIs and limit to 3 decimal places
  ) %>%
  select(Outcome = Model, `Point Estimate` = PE, `95% CI` = CI, `Unadjusted P-Value` = p_value)
tab_freq
```

```{r}
tab_freq_adh <- purrr::map_df(mod_list, summarise_lm_beta2, .id = "Model") %>%
  mutate(
    Outcome = sub("_adh$", "", Model),
    adh = grepl("_adh$", Model),
    CI = sprintf("(%0.3f, %0.3f)", CI_2.5, CI_97.5)
  ) %>%
  select(Outcome, adh, PE, CI, p_value) %>%
  pivot_wider(
    names_from = adh,
    values_from = c(PE, CI, p_value),
    names_prefix = "",
    names_sep = "_"
  ) %>%
  rename(
    PE = PE_FALSE, 
    PE_adh = PE_TRUE, 
    CI = CI_FALSE, 
    CI_adh = CI_TRUE, 
    p_value = p_value_FALSE, 
    p_value_adh = p_value_TRUE
  )  %>%
  mutate(
    Percent_Change = 100 * (PE_adh - PE) / abs(PE)
  ) %>%
  select(Outcome, `Point Estimate` = PE_adh, `Change in PE (%)` = Percent_Change, `95% CI` = CI_adh, `Unadjusted P-Value` = p_value_adh)
tab_freq_adh
```

# Bayesian analysis

## Set up STAN - only run one time, then comment out
```{r}
# # Dependencies
# library(cmdstanr)
# library(bayesplot)  # diagnostic plots of the MCMC chains
# library(posterior)  # for summarizing posterior draws
# library(bayestestR) # for calculating higest density posterior intervals
# library(mcmcse)     # for calculating MCMCSE's
# library(loo)        # for getting model fit statistics (WAIC and LOO-IC)
# library(dplyr)
# library(tibble)
# 
# ####Bayesian Analysis
# ###########################################################################
# # STEP 1: Define the model
# # This is a general linear regression set up for STAN
# # It is using a half normal prior on sigma
# # and normal priors on the regression coefficients
# # once the stan file is written it can be reused
# 
# # Do not need to rerun after initial. Works for any
# # linear regression you would like to do. Uses half normal
# # Saves program in correct file format in STAN
# ###########################################################################
# 
# stan_file <- write_stan_file("data {
#   int<lower=0> N;                  // number of observations
#   int<lower=0> P;                  // number of predictors including intercept
#   matrix[N, P] X;                  // design matrix (first column = intercept)
#   vector[N] y;                     // outcome
# 
#   vector[P] prior_mean;            // prior means for each beta
#   vector<lower=0>[P] prior_sd;     // prior SDs for each beta
# 
#   real<lower=0> sigma_prior_sd;    // SD for half-normal prior on sigma
# }
# 
# parameters {
#   vector[P] beta;                  // regression coefficients
#   real<lower=0> sigma;             // residual SD
# }
# 
# model {
#   // Vectorized priors for regression coefficients
#   beta ~ normal(prior_mean, prior_sd);
# 
#   // Half-normal prior for sigma
#   sigma ~ normal(0, sigma_prior_sd);
# 
#   // Likelihood
#   y ~ normal(X * beta, sigma);
# 
# 
# }
# 
# generated quantities {
#   // log likelihood for each observation for calculating model fit stats
#   vector[N] log_lik;
#   for (n in 1:N) {
#     log_lik[n] = normal_lpdf(y[n] | X[n] * beta, sigma);}
#   }", dir="STAN", basename='linear_regression_half_normal'
# )
# 
# ###########################################################################
# # STEP 2: Compile the Stan program
# ###########################################################################
mod <- cmdstan_model('STAN/linear_regression_half_normal.stan')
```

## Run models

Set up data
```{r}
# Outcome
y1 <- dat_wide$log10VLOAD_2
y2 <- dat_wide$LEU3N_2
y3 <- dat_wide$AGG_PHYS_2
y4 <- dat_wide$AGG_MENT_2


# Design matrix
X1a <- model.matrix(~ hard_drugs_0 + log10VLOAD_0 + age_0 + BMI_0 +
                    smoke_group_0 + educ_group_0 + race_group_0, data = dat_wide)
X1b <- model.matrix(~ hard_drugs_0 + log10VLOAD_0 + age_0 + BMI_0 +
                    smoke_group_0 + educ_group_0 + race_group_0 + adh_group_2,
                   data = dat_wide)
X2a <- model.matrix(~ hard_drugs_0 + LEU3N_0 + age_0 + BMI_0 +
                    smoke_group_0 + educ_group_0 + race_group_0, data = dat_wide)
X2b <- model.matrix(~ hard_drugs_0 + LEU3N_0 + age_0 + BMI_0 +
                    smoke_group_0 + educ_group_0 + race_group_0 + adh_group_2,
                   data = dat_wide)
X3a <- model.matrix(~ hard_drugs_0 + AGG_PHYS_0 + age_0 + BMI_0 +
                    smoke_group_0 + educ_group_0 + race_group_0, data = dat_wide)
X3b <- model.matrix(~ hard_drugs_0 + AGG_PHYS_0 + age_0 + BMI_0 +
                    smoke_group_0 + educ_group_0 + race_group_0 + adh_group_2,
                   data = dat_wide)
X4a <- model.matrix(~ hard_drugs_0 + AGG_MENT_0 + age_0 + BMI_0 +
                    smoke_group_0 + educ_group_0 + race_group_0, data = dat_wide)
X4b <- model.matrix(~ hard_drugs_0 + AGG_MENT_0 + age_0 + BMI_0 +
                    smoke_group_0 + educ_group_0 + race_group_0 + adh_group_2,
                   data = dat_wide)

N1a <- nrow(X1a)
P1a <- ncol(X1a)
N1b <- nrow(X1b)
P1b <- ncol(X1b)
N2a <- nrow(X2a)
P2a <- ncol(X2a)
N2b <- nrow(X2b)
P2b <- ncol(X2b)
N3a <- nrow(X3a)
P3a <- ncol(X3a)
N3b <- nrow(X3b)
P3b <- ncol(X3b)
N4a <- nrow(X4a)
P4a <- ncol(X4a)
N4b <- nrow(X4b)
P4b <- ncol(X4b)

# Priors: Normal(0, 1000) for all betas
prior_mean_1a <- rep(0, P1a); prior_sd_1a   <- rep(1000, P1a)
prior_mean_1b <- rep(0, P1b); prior_sd_1b   <- rep(1000, P1b)
prior_mean_2a <- rep(0, P2a); prior_sd_2a   <- rep(1000, P2a)
prior_mean_2b <- rep(0, P2b); prior_sd_2b   <- rep(1000, P2b)
prior_mean_3a <- rep(0, P3a); prior_sd_3a   <- rep(1000, P3a)
prior_mean_3b <- rep(0, P3b); prior_sd_3b   <- rep(1000, P3b)
prior_mean_4a <- rep(0, P4a); prior_sd_4a   <- rep(1000, P4a)
prior_mean_4b <- rep(0, P4b); prior_sd_4b   <- rep(1000, P4b)

# Half-normal(0, 1000) for sigma
sigma_prior_sd <- 1000

data_list_1a <- list(
  N = N1a, P = P1a, X = X1a, y = y1,
  prior_mean = prior_mean_1a,
  prior_sd = prior_sd_1a,
  sigma_prior_sd = sigma_prior_sd
)
data_list_1b <- list(
  N = N1b, P = P1b, X = X1b, y = y1,
  prior_mean = prior_mean_1b,
  prior_sd = prior_sd_1b,
  sigma_prior_sd = sigma_prior_sd
)
data_list_2a <- list(
  N = N2a, P = P2a, X = X2a, y = y2,
  prior_mean = prior_mean_2a,
  prior_sd = prior_sd_2a,
  sigma_prior_sd = sigma_prior_sd
)
data_list_2b <- list(
  N = N2b, P = P2b, X = X2b, y = y2,
  prior_mean = prior_mean_2b,
  prior_sd = prior_sd_2b,
  sigma_prior_sd = sigma_prior_sd
)
data_list_3a <- list(
  N = N3a, P = P3a, X = X3a, y = y3,
  prior_mean = prior_mean_3a,
  prior_sd = prior_sd_3a,
  sigma_prior_sd = sigma_prior_sd
)
data_list_3b <- list(
  N = N3b, P = P3b, X = X3b, y = y3,
  prior_mean = prior_mean_3b,
  prior_sd = prior_sd_3b,
  sigma_prior_sd = sigma_prior_sd
)
data_list_4a <- list(
  N = N4a, P = P4a, X = X4a, y = y4,
  prior_mean = prior_mean_4a,
  prior_sd = prior_sd_4a,
  sigma_prior_sd = sigma_prior_sd
)
data_list_4b <- list(
  N = N4b, P = P4b, X = X4b, y = y4,
  prior_mean = prior_mean_4b,
  prior_sd = prior_sd_4b,
  sigma_prior_sd = sigma_prior_sd
)
```

Fit model
```{r}
fit_vload <- mod$sample(data = data_list_1a, chains = 4, iter_warmup = 2000, iter_sampling = 5000, seed = 123)
fit_vload_adh <- mod$sample(data = data_list_1b, chains = 4, iter_warmup = 2000, iter_sampling = 5000, seed = 123)
fit_cd4 <- mod$sample(data = data_list_2a, chains = 4, iter_warmup = 2000, iter_sampling = 5000, seed = 123)
fit_cd4_adh <- mod$sample(data = data_list_2b, chains = 4, iter_warmup = 2000, iter_sampling = 5000, seed = 123)
fit_phys <- mod$sample(data = data_list_3a, chains = 4, iter_warmup = 2000, iter_sampling = 5000, seed = 123)
fit_phys_adh <- mod$sample(data = data_list_3b, chains = 4, iter_warmup = 2000, iter_sampling = 5000, seed = 123)
fit_ment <- mod$sample(data = data_list_4a, chains = 4, iter_warmup = 2000, iter_sampling = 5000, seed = 123)
fit_ment_adh <- mod$sample(data = data_list_4b, chains = 4, iter_warmup = 2000, iter_sampling = 5000, seed = 123)
```

## Model results

```{r}
fit_vload
fit_vload_adh
```
```{r}
fit_cd4
fit_cd4_adh
```
```{r}
fit_phys
fit_phys_adh
```
```{r}
fit_ment
fit_ment_adh
```


```{r}
summarise_beta2 <- function(fit, threshold) {
  
  draws <- fit$draws()
  draws_mat <- as_draws_matrix(draws)
  
  vals <- as.numeric(draws_mat[, "beta[2]"])
    # Extract outcome only   
  hpd <- hdi(vals, ci = 0.95)
    # HPDI
  
  tibble(
    Parameter = "beta[2]",
    PM = mean(vals),
    HPDI_2.5 = hpd[1],
    HPDI_97.5 = hpd[2],
    PP = mean(abs(vals) > threshold)
  )
}

model_list <- list(
  `Viral Load (log10 HIV copies/mL blood)` = fit_vload,
  `Viral Load (log10 HIV copies/mL blood)_adh` = fit_vload_adh,
  `CD4+ T Cell Count` = fit_cd4,
  `CD4+ T Cell Count_adh` = fit_cd4_adh,
  `Physical Quality of Life (SF-36 Aggregate)` = fit_phys,
  `Physical Quality of Life (SF-36 Aggregate)_adh` = fit_phys_adh,
  `Mental Quality of Life (SF-36 Aggregate)` = fit_ment,
  `Mental Quality of Life (SF-36 Aggregate)_adh` = fit_ment_adh
)

thresholds <- c(0.5, 0.5, 50, 50, 2, 2, 2, 2)
  # Clinically meaningful thresholds
tab_bayes <- purrr::map2_df(model_list, thresholds, summarise_beta2, .id = "Model") %>%
  mutate(
    HPDI = sprintf("(%0.3f, %0.3f)", HPDI_2.5, HPDI_97.5)
      # Combine HDPIs and limit to 3 decimal places
  ) %>%
  select(Outcome = Model, `Posterior Mean` = PM, `95% HPDI` = HPDI, `Posterior Probability` = PP)
tab_bayes
```
```{r}
tab_bayes_adh <- purrr::map2_df(model_list, thresholds, summarise_beta2, .id = "Model") %>%
  mutate(
    Outcome = sub("_adh$", "", Model),
    adh     = grepl("_adh$", Model),
    HPDI    = sprintf("(%0.3f, %0.3f)", HPDI_2.5, HPDI_97.5)
  ) %>%
  select(Outcome, adh, PM, HPDI, PP) %>%
  pivot_wider(
    names_from  = adh,
    values_from = c(PM, HPDI, PP)
  ) %>%
  rename(
    PM        = PM_FALSE,
    PM_adh    = PM_TRUE,
    HPDI      = HPDI_FALSE,
    HPDI_adh  = HPDI_TRUE,
    PP        = PP_FALSE,
    PP_adh    = PP_TRUE
  ) %>%
  mutate(
    Percent_Change = 100 * (PM_adh - PM) / abs(PM)
  ) %>%
  select(Outcome, `Posterior Mean` = PM_adh, `Change in PM (%)` = Percent_Change, `95% HPDI` = HPDI_adh, `Posterior Probability` = PP_adh)

tab_bayes_adh
```


# Combined table

```{r}
tab_comb <- dplyr::left_join(tab_freq, tab_bayes, by = "Outcome") %>%
  filter(!grepl("_adh$", Outcome)) %>%
  gt() %>%
  tab_spanner(
    label = "Frequentist",
    columns = c(`Point Estimate`, `95% CI`, `Unadjusted P-Value`)
  ) %>%
  tab_spanner(
    label = "Bayesian",
    columns = c(`Posterior Mean`, `95% HPDI`, `Posterior Probability`)
  ) %>%
  fmt_number(
    columns = c(`Point Estimate`, `Unadjusted P-Value`, `Posterior Mean`, `Posterior Probability`),
    decimals = 3
  ) %>%
  tab_header(
    title = "Comparison of Frequentist and Bayesian Estimates for Associations Between Hard Drug Use and Clinical Outcomes"
  ) %>%
  tab_footnote(
    footnote = "Comparison of Frequentist and Bayesian regression estimates for the association between baseline hard drug use and year‑2 outcomes. All models adjust for baseline outcome values, age, BMI, smoking status, education, and race. Bayesian models use weakly informative Normal(0, 1000) priors for regression coefficients and half‑Normal(0, 1000) priors for residual SD." 
  )
tab_comb
```

# Adherence table

```{r}
tab_adh <- dplyr::left_join(tab_freq_adh, tab_bayes_adh, by = "Outcome") %>%
  filter(!grepl("_adh$", Outcome)) %>%
  gt() %>%
  tab_spanner(
    label = "Frequentist",
    columns = c(`Point Estimate`, `Change in PE (%)`, `95% CI`, `Unadjusted P-Value`)
  ) %>%
  tab_spanner(
    label = "Bayesian",
    columns = c(`Posterior Mean`, `Change in PM (%)`, `95% HPDI`, `Posterior Probability`)
  ) %>%
  fmt_number(
    columns = c(`Point Estimate`, `Change in PE (%)`, `Unadjusted P-Value`, `Posterior Mean`, `Change in PM (%)`, `Posterior Probability`),
    decimals = 3
  ) %>%
  tab_header(
    title = "Impact of Adherence Adjustment on Frequentist and Bayesian Estimates for Associations Between Hard Drug Use and Clinical Outcomes"
  ) %>%
  tab_footnote(
    footnote = "Frequentist and Bayesian regression estimates for the association between baseline hard drug use and year‑2 outcomes after adjusting for adherence. All models also adjust for baseline outcome values, age, BMI, smoking status, education, and race. Percent change reflects the relative difference between adherence‑adjusted and non‑adjusted models, calculated for both the frequentist point estimate and the Bayesian posterior mean. Bayesian models use weakly informative Normal(0, 1000) priors for regression coefficients and half‑Normal(0, 1000) priors for residual SD." 
  )
tab_adh
```

# Figure

```{r}
dat_plot <- dat_wide %>%
  select(
    hard_drugs_0,
    log10VLOAD_2,
    LEU3N_2,
    AGG_PHYS_2,
    AGG_MENT_2
  ) %>%
  pivot_longer(
    cols = -hard_drugs_0,
    names_to = "Outcome",
    values_to = "Value"
  ) %>%
  mutate(
    hard_drugs_0 = factor(hard_drugs_0, labels = c("No Hard Drugs", "Hard Drugs")),
    Outcome = recode(
      Outcome,
      log10VLOAD_2 = "Viral Load (log10 HIV copies/mL blood)",
      LEU3N_2      = "CD4+ T Cell Count",
      AGG_PHYS_2   = "Physical Quality of Life (SF-36 Aggregate)",
      AGG_MENT_2   = "Mental Quality of Life (SF-36 Aggregate)"
    )
  )
freq_est <- purrr::map_df(mod_list, summarise_lm_beta2, .id = "Model") %>%
  filter(!grepl("_adh$", Model)) %>%
  transmute(
    Outcome = Model,
    Method = "Frequentist",
    Adjustment = "Unadjusted",
    Estimate = PE,
    lower = CI_2.5,
    upper = CI_97.5
  )
bayes_est <- purrr::map2_df(model_list, thresholds, summarise_beta2, .id = "Model") %>%
  filter(!grepl("_adh$", Model)) %>%
  transmute(
    Outcome = Model,
    Method = "Bayesian",
    Adjustment = "Unadjusted",
    Estimate = PM,
    lower = HPDI_2.5,
    upper = HPDI_97.5
  )
freq_est_adh <- purrr::map_df(mod_list, summarise_lm_beta2, .id = "Model") %>%
  filter(grepl("_adh$", Model)) %>%
  mutate(
    Outcome = sub("_adh$", "", Model),
    Method = "Frequentist",
    Adjustment = "Adherence-adjusted"
  ) %>%
  transmute(
    Outcome,
    Method,
    Adjustment,
    Estimate = PE,
    lower = CI_2.5,
    upper = CI_97.5
  )
bayes_est_adh <- purrr::map2_df(model_list, thresholds, summarise_beta2, .id = "Model") %>%
  filter(grepl("_adh$", Model)) %>%
  mutate(
    Outcome = sub("_adh$", "", Model),
    Method = "Bayesian",
    Adjustment = "Adherence-adjusted"
  ) %>%
  transmute(
    Outcome,
    Method,
    Adjustment,
    Estimate = PM,
    lower = HPDI_2.5,
    upper = HPDI_97.5
  )

#est_df <- bind_rows(freq_est, bayes_est)
dat_est <- bind_rows(
  freq_est,
  bayes_est,
  freq_est_adh,
  bayes_est_adh
) %>%
  left_join(
    dat_plot %>% group_by(Outcome, hard_drugs_0) %>% summarize(mean_val = mean(Value)),
    by = "Outcome"
  ) %>%
  mutate(
    est_y = mean_val + Estimate
  )
ggplot() +
  # Raw data points
  geom_point(
    data = dat_plot,
    aes(x = hard_drugs_0, y = Value),
    alpha = 0.25,
    color = "gray40",
    position = position_jitter(width = 0.15, height = 0)
  ) +
  
  # Model estimates (PE/PM)
  geom_point(
    data = dat_est,
    aes(x = hard_drugs_0, y = est_y, color = Method, shape = Adjustment),
    size = 3,
    position = position_dodge(width = 0.4)
  ) +
  
  # Error bars (CI / HPDI)
  geom_errorbar(
    data = dat_est,
    aes(
      x = hard_drugs_0,
      ymin = mean_val + lower,
      ymax = mean_val + upper,
      color = Method
    ),
    width = 0.15,
    position = position_dodge(width = 0.4)
  ) +
  facet_wrap(~ Outcome, scales = "free_y") +
  scale_shape_manual(values = c("Unadjusted" = 16, "Adherence-adjusted" = 17)) +
  labs(
    x = "Hard Drug Use at Baseline",
    y = "Outcome Value",
    title = "Observed Outcomes with Frequentist and Bayesian Estimated Effects Overlaid",
    color = "Model Type",
    caption = "Observed second year outcomes by baseline hard drug use with overlaid Frequentist and Bayesian estimates of the adjusted effect of hard drug usage, separated by unadjusted estimates and estimates that accounted for adherence. Grey points show individual observations, colored points/ranges show estimated effects and 95% CI/HPDI. Models adjust for baseline outcome values, age, BMI, smoking status, education, and race, with Bayesian models using weakly informative Normal(0, 1000) priors for coefficients and half Normal(0, 1000) priors for residual SD."
  ) +
  theme_bw() +
  theme(
    strip.text = element_text(size = 11),
    legend.position = "bottom",
    plot.caption = ggtext::element_textbox_simple(
    size = 9,
    width = unit(1, "npc"),   # allow full-width wrapping
    margin = margin(t = 10),
    lineheight = 1.1
    ),
    plot.margin = margin(10, 10, 40, 10)
  )

```

```{r}
dat_plot <- dat_wide %>%
  select(
    hard_drugs_0,
    log10VLOAD_2,
    LEU3N_2,
    AGG_PHYS_2,
    AGG_MENT_2
  ) %>%
  pivot_longer(
    cols = -hard_drugs_0,
    names_to = "Outcome",
    values_to = "Value"
  ) %>%
  mutate(
    hard_drugs_0 = factor(hard_drugs_0, labels = c("No Hard Drugs", "Hard Drugs")),
    Outcome = recode(
      Outcome,
      log10VLOAD_2 = "Viral Load (log10 HIV copies/mL blood)",
      LEU3N_2      = "CD4+ T Cell Count",
      AGG_PHYS_2   = "Physical Quality of Life (SF-36 Aggregate)",
      AGG_MENT_2   = "Mental Quality of Life (SF-36 Aggregate)"
    )
  )

ggplot(dat_plot, aes(x = hard_drugs_0, y = Value)) +
  geom_violin(fill = "gray85", color = "gray60") +
  geom_jitter(alpha = 0.25, width = 0.15, color = "gray40") +
  facet_wrap(~ Outcome, scales = "free_y") +
  labs(
    x = "Hard Drug Use at Baseline",
    y = "Outcome Value",
    title = "Observed Year‑2 Outcomes by Baseline Hard Drug Use"
  ) +
  theme_bw() +
  theme(
    strip.text = element_text(size = 11),
    legend.position = "none"
  )
```

```{r}
# Prepare data for forest plot
forest_df <- purrr::map_df(mod_list, summarise_lm_beta2, .id = "Model") %>%
  transmute(
    Outcome = Model,
    Method = "Frequentist",
    Estimate = PE,
    lower = CI_2.5,
    upper = CI_97.5
  ) %>%
  bind_rows(
    purrr::map2_df(model_list, thresholds, summarise_beta2, .id = "Model") %>%
      transmute(
        Outcome = Model,
        Method = "Bayesian",
        Estimate = PM,
        lower = HPDI_2.5,
        upper = HPDI_97.5
      )
  ) %>%
  mutate(
    Outcome = recode(
      Outcome,
      log10VLOAD_2 = "Viral Load (log10 HIV copies/mL blood)",
      LEU3N_2      = "CD4+ T Cell Count",
      AGG_PHYS_2   = "Physical Quality of Life (SF-36 Aggregate)",
      AGG_MENT_2   = "Mental Quality of Life (SF-36 Aggregate)"
    )
  ) %>%
  mutate(
    cm_diff = case_when(
      Outcome == "Viral Load (log10 HIV copies/mL blood)" ~ 0.5,
      Outcome == "CD4+ T Cell Count" ~ 50,
      Outcome == "Physical Quality of Life (SF-36 Aggregate)" ~ 2,
      Outcome == "Mental Quality of Life (SF-36 Aggregate)" ~ 2
    )
  )

ggplot(forest_df, aes(x = Estimate, y = Method, color = Method)) +
  geom_point(size = 3) +
  geom_errorbar(aes(xmin = lower, xmax = upper), width = 0.2) +
  facet_wrap(~ Outcome, scales = "free_x") +
  geom_vline(xintercept = 0, linetype = "dashed") +
  labs(
    x = "Adjusted Effect of Baseline Hard Drug Use",
    y = "",
    title = "Adjusted Associations Between Baseline Hard Drug Use and Year‑2 Outcomes"
  ) +
  theme_bw() +
  theme(
    legend.position = "bottom",
    strip.text = element_text(size = 11)
  ) + 
  geom_rect(
    aes(
      xmin = -cm_diff, xmax = cm_diff,
      ymin = -Inf, ymax = Inf
    ),
    fill = "grey90", alpha = 0.3,
    inherit.aes = FALSE
  )
```

```{r}
# ---- Prepare data for forest plot with adherence adjustment ----

forest_df <-
  # Frequentist unadjusted
  purrr::map_df(mod_list, summarise_lm_beta2, .id = "Model") %>%
  mutate(
    Adjustment = ifelse(grepl("_adh$", Model), "Adherence-adjusted", "Unadjusted"),
    Outcome = sub("_adh$", "", Model),
    Method = "Frequentist"
  ) %>%
  transmute(
    Outcome,
    Method,
    Adjustment,
    Estimate = PE,
    lower = CI_2.5,
    upper = CI_97.5
  ) %>%
  
  # Add Bayesian
  bind_rows(
    purrr::map2_df(model_list, thresholds, summarise_beta2, .id = "Model") %>%
      mutate(
        Adjustment = ifelse(grepl("_adh$", Model), "Adherence-adjusted", "Unadjusted"),
        Outcome = sub("_adh$", "", Model),
        Method = "Bayesian"
      ) %>%
      transmute(
        Outcome,
        Method,
        Adjustment,
        Estimate = PM,
        lower = HPDI_2.5,
        upper = HPDI_97.5
      )
  ) %>%
  
  # Recode outcome names + add clinically meaningful thresholds
  mutate(
    Outcome = recode(
      Outcome,
      log10VLOAD_2 = "Viral Load (log10 HIV copies/mL blood)",
      LEU3N_2      = "CD4+ T Cell Count",
      AGG_PHYS_2   = "Physical Quality of Life (SF-36 Aggregate)",
      AGG_MENT_2   = "Mental Quality of Life (SF-36 Aggregate)"
    ),
    cm_diff = case_when(
      Outcome == "Viral Load (log10 HIV copies/mL blood)" ~ 0.5,
      Outcome == "CD4+ T Cell Count" ~ 50,
      Outcome == "Physical Quality of Life (SF-36 Aggregate)" ~ 2,
      Outcome == "Mental Quality of Life (SF-36 Aggregate)" ~ 2
    ),
    # ---- Set facet order here ----
    Outcome = factor(
      Outcome,
      levels = c(
        "Viral Load (log10 HIV copies/mL blood)",
        "CD4+ T Cell Count",
        "Physical Quality of Life (SF-36 Aggregate)",
        "Mental Quality of Life (SF-36 Aggregate)"
      )
    )
  )

# ---- Plot ----

p <- ggplot(forest_df, aes(x = Estimate, y = Method, color = Method, shape = Adjustment)) +
  
  geom_rect(
    aes(
      xmin = -cm_diff, xmax = cm_diff,
      ymin = -Inf, ymax = Inf
    ),
    fill = "grey90", alpha = 0.3,
    inherit.aes = FALSE
  ) +
  
  geom_vline(xintercept = 0, linetype = "dashed") +
  
  geom_point(size = 3, position = position_dodge(width = 0.5)) +
  
  geom_errorbar(
    aes(xmin = lower, xmax = upper),
    width = 0.2,
    position = position_dodge(width = 0.5)
  ) +
  
  facet_wrap(~ Outcome, scales = "free_x") +
  
  scale_shape_manual(values = c(
    "Unadjusted" = 16,
    "Adherence-adjusted" = 17
  )) +
  
  labs(
    x = "Adjusted Effect of Baseline Hard Drug Use",
    y = "",
    title = "Adjusted Associations Between Baseline Hard Drug Use and 2nd Year Outcomes",
    color = "Model Type",
    shape = "Adjustment",
  ) +
  
  theme_bw() +
  theme(
    legend.position = "bottom",
    strip.text = element_text(size = 11),
  )

library(cowplot)

caption_text <- "Adjusted associations between baseline hard drug use and 2nd year clinical outcomes estimated using Frequentist and Bayesian regression models, shown separately for models with and without adherence. All models adjust for baseline outcome values, age, BMI, smoking status, education, and race. Bayesian models use weakly informative Normal(0, 1000) priors for regression coefficients and half‑Normal(0, 1000) priors for residual standard deviations. Points represent the estimated effect of baseline hard drug use, with error bars showing 95% confidence intervals (Frequentist) or 95% highest posterior density intervals (Bayesian). The grey shaded region in each panel denotes the range corresponding to the two-sided clinically meaningful difference for that outcome (0.5 log10 for viral load, 50 cells for CD4 count, and 2 points for quality of life scales), providing a visual benchmark for interpreting the magnitude of estimated effects."

library(ggtext)
library(patchwork)

caption_panel <- ggplot() +
  annotate(
    "richtext",
    x = 0, y = 1,
    label = caption_text,
    hjust = 0,
    vjust = 1,
    size = 3,
    width = unit(0.95, "npc")
  ) +
  theme_void()

p / caption_panel + plot_layout(heights = c(1, 0.25))
```
```{r fig.cap="Adjusted associations between baseline hard drug use and 2nd year clinical outcomes estimated using Frequentist and Bayesian regression models, shown separately for models with and without adherence. All models adjust for baseline outcome values, age, BMI, smoking status, education, and race. Bayesian models use weakly informative Normal(0, 1000) priors for regression coefficients and half‑Normal(0, 1000) priors for residual standard deviations. Points represent the estimated effect of baseline hard drug use, with error bars showing 95% confidence intervals (Frequentist) or 95% highest posterior density intervals (Bayesian). The grey shaded region in each panel denotes the range corresponding to the two-sided clinically meaningful difference for that outcome (0.5 log10 for viral load, 50 cells for CD4 count, and 2 points for quality of life scales), providing a visual benchmark for interpreting the magnitude of estimated effects."}
p
```

