---
title: "P1_Draft_Analysis"
output: html_document
date: "2026-02-25"
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
library(dplyr)
library(tidyr)
library(purrr)
library(ggplot2)
library(table1)
dat_raw <- read.csv(here::here("P1/DataRaw", "hiv_6624_final.csv"))
```

```{r}
dat_sub <- dat_raw %>%
  filter(years %in% c(0, 2)) %>%
    # Filter to baseline and year 2
  mutate(
    race_group = case_when(
      RACE == 1 ~ "White nonHisp",
      RACE %in% c(2, 3, 4, 5, 6, 7, 8) ~ "Other",
      TRUE ~ NA_character_
    )
  ) %>%
  mutate(
    educ_group = case_when(
      EDUCBAS %in% c(1, 2, 3, 4) ~ "No College Degree",
      EDUCBAS %in% c(5, 6, 7) ~ "College Degree and above",
      TRUE ~ NA_character_
    )
  ) %>%
  mutate(
    smoke_group = case_when(
      SMOKE %in% c(1, 2) ~ "Non or Former Smoker",
      SMOKE == 3 ~ "Current Smoker",
      TRUE ~ NA_character_
    )
  )

baseline_drugs <- dat_sub %>%
  filter(years == 0) %>%
  select(newid, hard_drugs)
    # Identify baseline drug use by id
dat_sub <- dat_sub %>%
  left_join(baseline_drugs, by = "newid", suffix = c("", "_baseline"))
    # Add baseline drug use to dataset
dat_vload <- dat_sub %>%
  group_by(newid) %>%
  filter(
    any(years == 0 & !is.na(VLOAD)) &
    any(years == 2 & !is.na(VLOAD))
  ) %>%
  ungroup()

dat_leu3n <- dat_sub %>%
  group_by(newid) %>%
  filter(
    any(years == 0 & !is.na(LEU3N)) &
    any(years == 2 & !is.na(LEU3N))
  ) %>%
  ungroup()

dat_phys <- dat_sub %>%
  group_by(newid) %>%
  filter(
    any(years == 0 & !is.na(AGG_PHYS)) &
    any(years == 2 & !is.na(AGG_PHYS))
  ) %>%
  ungroup()

dat_ment <- dat_sub %>%
  group_by(newid) %>%
  filter(
    any(years == 0 & !is.na(AGG_MENT)) &
    any(years == 2 & !is.na(AGG_MENT))
  ) %>%
  ungroup()

dat_sub_outcomes <- dat_sub[dat_sub$newid %in% Reduce(intersect, list(dat_vload$newid, dat_leu3n$newid, dat_phys$newid, dat_ment$newid)),]
  # Subset to those that have all four outcomes at times 0 and 2
dat_sub_outcomes <- dat_sub_outcomes %>%
  group_by(newid) %>%
  filter(
    any(years == 0 & !is.na(hard_drugs))
  ) %>%
  ungroup() %>% # Subset to those that also have hard drug use at time 0
  mutate(
    hard_drugs_baseline = factor(hard_drugs_baseline, levels = c(0, 1), labels = c("No Hard Drug Use", "Hard Drug Use"))
  ) %>%
  mutate(BMI = if_else(BMI < 0 | BMI > 250, NA_real_, BMI))
    # Exclude for being clinically implausible (999 and -1 are insufficient or improbable data, also some values in the 500s)

table1a <- table1(~ log10(VLOAD) + LEU3N + AGG_PHYS + AGG_MENT + age + BMI + smoke_group + educ_group + race_group | hard_drugs_baseline, dat_sub_outcomes %>% filter(years == 0), caption="Baseline Characteristics by Baseline Hard Drug Use")
table1b <- table1(~ log10(VLOAD) + LEU3N + AGG_PHYS + AGG_MENT + ADH | hard_drugs_baseline, dat_sub_outcomes %>% filter(years == 2), caption="Year 2 Characteristics by Baseline Hard Drug Use")

dat_sub_base <- dat_sub_outcomes %>%
  filter(years == 0,
         if_all(c(age, BMI, smoke_group, educ_group, race_group), ~ !is.na(.x)))
```

TODO: analysis steps
- collapse ADH to upper 2 and lower 2 groups (>95% and <95%)
- reduce to 463 individuals with complete covariate data
- use dplyr to transform dat_sub so that you have y2 log10VLOAD/other outcomes and baseline hard drug use/covariates/y2 adherence all available in one single row
- model with simple linear regression with separate models for all 4 outcomes
  - store coefficients, unadjusted pvals, 95% CIs
  - model with adherence and store % change in outcome (are there differences in adherence/relationship?)
- get STAN to run and model with bayesian linear regression with N(0,1000) for regression coefficients and HN(0,1000) for error. 
  - store posterior mean, HDPI, posterior probability (use clinically meaningful differences)
  - model with adherence and store % change in outcome
- Make table of outcomes that combines frequentist and bayesian outcomes previously stored (table 3)
- Consider what figures or any additional tables that you want to include
  - perhaps separate table of adherence % changes to table 3
  - perhaps some figure to visualize significant outcomes by hard drug status/somehow visualize the models (think boxplot of each outcome with 2 for each color coded by hard drug use, each outcome as a panel)
- write report 
- make presentation


