---
title: "P1_Draft_Analysis"
output: html_document
date: "2026-02-25"
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
library(dplyr)
library(tidyr)
library(purrr)
library(ggplot2)
library(table1)
dat_raw <- read.csv(here::here("P1/DataRaw", "hiv_6624_final.csv"))
```

```{r}
dat_sub <- dat_raw %>%
  filter(
    years %in% c(0, 2)
  ) %>%
    # Filter to baseline and year 2
  mutate(
    race_group = case_when(
      RACE == 1 ~ "White nonHisp",
      RACE %in% 2:8 ~ "Other",
      TRUE ~ NA_character_
    ), # Collapse race
    educ_group = case_when(
      EDUCBAS %in% 1:4 ~ "No College Degree",
      EDUCBAS %in% 5:7 ~ "College Degree and above",
      TRUE ~ NA_character_
    ), # Collapse education
    smoke_group = case_when(
      SMOKE %in% c(1, 2) ~ "Non or Former Smoker",
      SMOKE == 3 ~ "Current Smoker",
      TRUE ~ NA_character_
    ), # Collapse smoking status
    adh_group = case_when(
      ADH %in% c(1, 2) ~ "95% and Above",
      ADH %in% c(3, 4) ~ "<95%",
      TRUE ~ NA_character_
    ), # Collapse adherence 
    hard_drugs = case_when(
      hard_drugs == 0 ~ "No",
      hard_drugs == 1 ~ "Yes",
      TRUE ~ NA_character_
    ),
      # Add labels to hard drugs
    BMI = if_else(
      BMI < 0 | BMI > 250, NA_real_, BMI
    ), # Exclude implausible BMI values
    log10VLOAD = log10(VLOAD)
      # Convert VLOAD to log10 scale
  )

dat_wide <- dat_sub %>%
  pivot_wider(
    id_cols = newid,
    names_from = years,
    values_from = c(
      log10VLOAD, LEU3N, AGG_PHYS, AGG_MENT,
      age, BMI, smoke_group, educ_group, race_group,
      hard_drugs, adh_group
    ),
    names_glue = "{.value}_{years}"
  )

required_outcomes <- c("log10VLOAD_0", "log10VLOAD_2", "LEU3N_0", "LEU3N_2",
                       "AGG_PHYS_0", "AGG_PHYS_2", "AGG_MENT_0", "AGG_MENT_2")
required_predictors <- c("age_0", "BMI_0", "smoke_group_0", "educ_group_0",
                         "race_group_0", "hard_drugs_0", "adh_group_2")

dat_wide <- dat_wide %>%
  filter(
    complete.cases(across(c(required_outcomes, required_predictors)))
  ) %>% # Filter to those without missing outcomes or predictors
  select(
    newid, all_of(required_outcomes), all_of(required_predictors)
  ) # Only select relevant columns
  # 463 final sample size

table1a <- table1(~ log10VLOAD_0 + LEU3N_0 + AGG_PHYS_0 + AGG_MENT_0 + age_0 + BMI_0 + smoke_group_0 + educ_group_0 + race_group_0 | hard_drugs_0, dat_wide, caption="Baseline Characteristics by Baseline Hard Drug Use")
table1b <- table1(~ log10VLOAD_2 + LEU3N_2 + AGG_PHYS_2 + AGG_MENT_2 + adh_group_2 | hard_drugs_0, dat_wide, caption="Year 2 Characteristics by Baseline Hard Drug Use")
```

TODO: analysis steps
- collapse ADH to upper 2 and lower 2 groups (>95% and <95%)
- reduce to 463 individuals with complete covariate data
- use dplyr to transform dat_sub so that you have y2 log10VLOAD/other outcomes and baseline hard drug use/covariates/y2 adherence all available in one single row
- model with simple linear regression with separate models for all 4 outcomes
------------------------------------------------------------------------------
  - store coefficients, unadjusted pvals, 95% CIs
  - model with adherence and store % change in outcome (are there differences in adherence/relationship?)
- get STAN to run and model with bayesian linear regression with N(0,1000) for regression coefficients and HN(0,1000) for error. 
  - store posterior mean, HDPI, posterior probability (use clinically meaningful differences)
  - model with adherence and store % change in outcome
- Make table of outcomes that combines frequentist and bayesian outcomes previously stored (table 3)
- Consider what figures or any additional tables that you want to include
  - perhaps separate table of adherence % changes to table 3
  - perhaps some figure to visualize significant outcomes by hard drug status/somehow visualize the models (think boxplot of each outcome with 2 for each color coded by hard drug use, each outcome as a panel)
- write report 
- make presentation

```{r}
mod_vload <- lm(log10VLOAD_2 ~ hard_drugs_0 + log10VLOAD_0 + age_0 + BMI_0 +
                  smoke_group_0 + educ_group_0 + race_group_0, data = dat_wide)
mod_cd4 <- lm(LEU3N_2 ~ hard_drugs_0 + LEU3N_0 + age_0 + BMI_0 + smoke_group_0 +
                educ_group_0 + race_group_0, data = dat_wide)
mod_phys <- lm(AGG_PHYS_2 ~ hard_drugs_0 + AGG_PHYS_0 + age_0 + BMI_0 +
                 smoke_group_0 + educ_group_0 + race_group_0, data = dat_wide)
mod_ment <- lm(AGG_MENT_2 ~ hard_drugs_0 + AGG_MENT_0 + age_0 + BMI_0 +
                 smoke_group_0 + educ_group_0 + race_group_0, data = dat_wide)
```

```{r}
mod_vload_adh <- lm(log10VLOAD_2 ~ hard_drugs_0 + log10VLOAD_0 + age_0 + BMI_0 +
                      smoke_group_0 + educ_group_0 + race_group_0 + adh_group_2,
                    data = dat_wide)
mod_cd4_adh <- lm(LEU3N_2 ~ hard_drugs_0 + LEU3N_0 + age_0 + BMI_0 + 
                    smoke_group_0 + educ_group_0 + race_group_0 + adh_group_2,
                  data = dat_wide)
mod_phys_adh <- lm(AGG_PHYS_2 ~ hard_drugs_0 + AGG_PHYS_0 + age_0 + BMI_0 +
                     smoke_group_0 + educ_group_0 + race_group_0 + adh_group_2,
                   data = dat_wide)
mod_ment_adh <- lm(AGG_MENT_2 ~ hard_drugs_0 + AGG_MENT_0 + age_0 + BMI_0 +
                     smoke_group_0 + educ_group_0 + race_group_0 + adh_group_2,
                   data = dat_wide)
```

```{r}
summary(mod_vload)
summary(mod_vload_adh)
```
```{r}
summary(mod_cd4)
summary(mod_cd4_adh)
```
```{r}
summary(mod_phys)
summary(mod_phys_adh)
```
```{r}
summary(mod_ment)
summary(mod_ment_adh)
```

