---
output:
  bookdown::pdf_document2:
    toc: false
    number_sections: false
documentclass: article
geometry: margin=1in
fontsize: 12pt
linestretch: 2
fig.cap: true
header-includes:
  - \usepackage{float}
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
library(dplyr)
library(tidyr)
library(tidyverse)
library(purrr)
library(ggplot2)
library(table1)
library(cmdstanr)
library(posterior)
library(tidybayes)
library(tibble)
library(mcmcse)
library(gt)
dat_raw <- read.csv(here::here("P1/DataRaw", "hiv_6624_final.csv"))
```

<!-- Title here: -->
# \textit{Analysis Plan for HAART Outcomes by Drug Use}

<!-- Name here: -->
Ezra Weible[^1]

[^1]: Biostatistics Masters Student, Colorado School of Public Health, CU Anschutz Medical Campus

```{r data_management, echo=FALSE, message=FALSE, results='hide', warning=FALSE, cache=TRUE}
dat_sub <- dat_raw %>%
  filter(
    years %in% c(0, 2)
  ) %>%
    # Filter to baseline and year 2
  mutate(
    race_group = case_when(
      RACE == 1 ~ "White nonHisp",
      RACE %in% 2:8 ~ "Other",
      TRUE ~ NA_character_
    ), # Collapse race
    educ_group = case_when(
      EDUCBAS %in% 1:4 ~ "No College Degree",
      EDUCBAS %in% 5:7 ~ "College Degree and above",
      TRUE ~ NA_character_
    ), # Collapse education
    smoke_group = case_when(
      SMOKE %in% c(1, 2) ~ "Non or Former Smoker",
      SMOKE == 3 ~ "Current Smoker",
      TRUE ~ NA_character_
    ), # Collapse smoking status
    adh_group = case_when(
      ADH %in% c(1, 2) ~ "95% and Above",
      ADH %in% c(3, 4) ~ "<95%",
      TRUE ~ NA_character_
    ), # Collapse adherence 
    hard_drugs = case_when(
      hard_drugs == 0 ~ "No",
      hard_drugs == 1 ~ "Yes",
      TRUE ~ NA_character_
    ),
      # Add labels to hard drugs
    BMI = if_else(
      BMI < 0 | BMI > 250, NA_real_, BMI
    ), # Exclude implausible BMI values
    log10VLOAD = log10(VLOAD)
      # Convert VLOAD to log10 scale
  )

dat_wide <- dat_sub %>%
  pivot_wider(
    id_cols = newid,
    names_from = years,
    values_from = c(
      log10VLOAD, LEU3N, AGG_PHYS, AGG_MENT,
      age, BMI, smoke_group, educ_group, race_group,
      hard_drugs, adh_group
    ),
    names_glue = "{.value}_{years}"
  ) # Pivot to wide

required_outcomes <- c("log10VLOAD_0", "log10VLOAD_2", "LEU3N_0", "LEU3N_2",
                       "AGG_PHYS_0", "AGG_PHYS_2", "AGG_MENT_0", "AGG_MENT_2")
  # Store required outcomes
required_predictors <- c("age_0", "BMI_0", "smoke_group_0", "educ_group_0",
                         "race_group_0", "hard_drugs_0", "adh_group_2")
  # Store required predictors

dat_wide <- dat_wide %>%
  filter(
    complete.cases(across(c(required_outcomes, required_predictors)))
  ) %>% # Filter to those without missing outcomes or predictors
  select(
    newid, all_of(required_outcomes), all_of(required_predictors)
  ) # Only select relevant columns
      # 463 final sample size

table1 <- table1(~ log10VLOAD_0 + LEU3N_0 + AGG_PHYS_0 + AGG_MENT_0 + age_0 + BMI_0 + smoke_group_0 + educ_group_0 + race_group_0 | hard_drugs_0, dat_wide, caption="Baseline Characteristics by Baseline Hard Drug Use")
  # Generate table 1 (baseline)
table2 <- table1(~ log10VLOAD_2 + LEU3N_2 + AGG_PHYS_2 + AGG_MENT_2 + adh_group_2 | hard_drugs_0, dat_wide, caption="Year 2 Characteristics by Baseline Hard Drug Use")
  # Generate table 2 (year 2)
```
```{r frequentist, echo=FALSE, results='hide', message=FALSE, warning=FALSE, cache=TRUE}
## Frequentist models
mod_vload <- lm(log10VLOAD_2 ~ hard_drugs_0 + log10VLOAD_0 + age_0 + BMI_0 +
                  smoke_group_0 + educ_group_0 + race_group_0, data = dat_wide)
mod_cd4 <- lm(LEU3N_2 ~ hard_drugs_0 + LEU3N_0 + age_0 + BMI_0 + smoke_group_0 +
                educ_group_0 + race_group_0, data = dat_wide)
mod_phys <- lm(AGG_PHYS_2 ~ hard_drugs_0 + AGG_PHYS_0 + age_0 + BMI_0 +
                 smoke_group_0 + educ_group_0 + race_group_0, data = dat_wide)
mod_ment <- lm(AGG_MENT_2 ~ hard_drugs_0 + AGG_MENT_0 + age_0 + BMI_0 +
                 smoke_group_0 + educ_group_0 + race_group_0, data = dat_wide)

## Frequentist models with adherence
mod_vload_adh <- lm(log10VLOAD_2 ~ hard_drugs_0 + log10VLOAD_0 + age_0 + BMI_0 +
                      smoke_group_0 + educ_group_0 + race_group_0 + adh_group_2,
                    data = dat_wide)
mod_cd4_adh <- lm(LEU3N_2 ~ hard_drugs_0 + LEU3N_0 + age_0 + BMI_0 + 
                    smoke_group_0 + educ_group_0 + race_group_0 + adh_group_2,
                  data = dat_wide)
mod_phys_adh <- lm(AGG_PHYS_2 ~ hard_drugs_0 + AGG_PHYS_0 + age_0 + BMI_0 +
                     smoke_group_0 + educ_group_0 + race_group_0 + adh_group_2,
                   data = dat_wide)
mod_ment_adh <- lm(AGG_MENT_2 ~ hard_drugs_0 + AGG_MENT_0 + age_0 + BMI_0 +
                     smoke_group_0 + educ_group_0 + race_group_0 + adh_group_2,
                   data = dat_wide)

## View model outputs
summary(mod_vload)
summary(mod_vload_adh)
summary(mod_cd4)
summary(mod_cd4_adh)
summary(mod_phys)
summary(mod_phys_adh)
summary(mod_ment)
summary(mod_ment_adh)
```
```{r freq_tables, echo=FALSE, message=FALSE, results='hide', warning=FALSE, cache=TRUE}
summarize_lm_beta2 <- function(mod) { # Function to extract frequentist output
  sm <- summary(mod)
    # Full model summary
  coefs <- sm$coefficients
    # Coefficients table
  est <- coefs["hard_drugs_0Yes", "Estimate"]
    # Pull point estimate
  se  <- coefs["hard_drugs_0Yes", "Std. Error"]
    # Pull standard error
  p   <- coefs["hard_drugs_0Yes", "Pr(>|t|)"]
    # Pull p-value
  ci_low  <- est - 1.96 * se
  ci_high <- est + 1.96 * se
    # Calculate 95% CI
  
  tibble(
    Parameter = "hard_drugs_0Yes",
    PE = est,
    CI_2.5 = ci_low,
    CI_97.5 = ci_high,
    p_value = p
  ) # Return results as a tibble
}

mod_list <- list(
  `Viral Load (log10 HIV copies/mL blood)` = mod_vload,
  `Viral Load (log10 HIV copies/mL blood)_adh` = mod_vload_adh,
  `CD4+ T Cell Count` = mod_cd4,
  `CD4+ T Cell Count_adh` = mod_cd4_adh,
  `Physical Quality of Life (SF-36 Aggregate)` = mod_phys,
  `Physical Quality of Life (SF-36 Aggregate)_adh` = mod_phys_adh,
  `Mental Quality of Life (SF-36 Aggregate)` = mod_ment,
  `Mental Quality of Life (SF-36 Aggregate)_adh` = mod_ment_adh
) # List of frequentist models, renamed for tables

tab_freq <- purrr::map_df(mod_list, summarize_lm_beta2, .id = "Model") %>%
    # Apply function to each model
  mutate(
    CI = sprintf("(%0.3f, %0.3f)", CI_2.5, CI_97.5)
  ) %>% # Combine CIs and limit to 3 decimal places
  select(Outcome = Model, `Point Estimate` = PE, `95% CI` = CI, `Unadjusted P-Value` = p_value)
    # Select and rename final columns

tab_freq_adh <- purrr::map_df(mod_list, summarize_lm_beta2, .id = "Model") %>%
    # Apply function to each model
  mutate(
    Outcome = sub("_adh$", "", Model),
    adh = grepl("_adh$", Model),
      # Select only results from adherence models
    CI = sprintf("(%0.3f, %0.3f)", CI_2.5, CI_97.5)
      # Combine CIs and limit to 3 decimal places
  ) %>%
  select(Outcome, adh, PE, CI, p_value) %>%
    # Select relevant columns
  pivot_wider(
    names_from = adh,
    values_from = c(PE, CI, p_value),
    names_prefix = "",
    names_sep = "_"
  ) %>% # Pivot to wide and add subscript to note adh
  rename(
    PE = PE_FALSE, 
    PE_adh = PE_TRUE, 
    CI = CI_FALSE, 
    CI_adh = CI_TRUE, 
    p_value = p_value_FALSE, 
    p_value_adh = p_value_TRUE
  ) %>% # Rename columns 
  mutate(
    Percent_Change = 100 * (PE_adh - PE) / abs(PE)
  ) %>% # Calculate impact of adherence
  select(Outcome, `Point Estimate` = PE_adh, `Change in PE (%)` = Percent_Change, `95% CI` = CI_adh, `Unadjusted P-Value` = p_value_adh)
    # Select and rename final columns
```
```{r stan_setup, echo=FALSE, message=FALSE, results='hide', warning=FALSE, cache=TRUE}
## Set up STAN - only run one time, then comment out

# # Dependencies
# library(cmdstanr)
# library(bayesplot)  # diagnostic plots of the MCMC chains
# library(posterior)  # for summarizing posterior draws
# library(bayestestR) # for calculating higest density posterior intervals
# library(mcmcse)     # for calculating MCMCSE's
# library(loo)        # for getting model fit statistics (WAIC and LOO-IC)
# library(dplyr)
# library(tibble)
# 
# ####Bayesian Analysis
# ###########################################################################
# # STEP 1: Define the model
# # This is a general linear regression set up for STAN
# # It is using a half normal prior on sigma
# # and normal priors on the regression coefficients
# # once the stan file is written it can be reused
# 
# # Do not need to rerun after initial. Works for any
# # linear regression you would like to do. Uses half normal
# # Saves program in correct file format in STAN
# ###########################################################################
# 
# stan_file <- write_stan_file("data {
#   int<lower=0> N;                  // number of observations
#   int<lower=0> P;                  // number of predictors including intercept
#   matrix[N, P] X;                  // design matrix (first column = intercept)
#   vector[N] y;                     // outcome
# 
#   vector[P] prior_mean;            // prior means for each beta
#   vector<lower=0>[P] prior_sd;     // prior SDs for each beta
# 
#   real<lower=0> sigma_prior_sd;    // SD for half-normal prior on sigma
# }
# 
# parameters {
#   vector[P] beta;                  // regression coefficients
#   real<lower=0> sigma;             // residual SD
# }
# 
# model {
#   // Vectorized priors for regression coefficients
#   beta ~ normal(prior_mean, prior_sd);
# 
#   // Half-normal prior for sigma
#   sigma ~ normal(0, sigma_prior_sd);
# 
#   // Likelihood
#   y ~ normal(X * beta, sigma);
# 
# 
# }
# 
# generated quantities {
#   // log likelihood for each observation for calculating model fit stats
#   vector[N] log_lik;
#   for (n in 1:N) {
#     log_lik[n] = normal_lpdf(y[n] | X[n] * beta, sigma);}
#   }", dir="STAN", basename='linear_regression_half_normal'
# )
# 
# ###########################################################################
# # STEP 2: Compile the Stan program
# ###########################################################################
mod <- cmdstan_model('STAN/linear_regression_half_normal.stan')
```
```{r bayes_setup, echo=FALSE, message=FALSE, results='hide', warning=FALSE, cache=TRUE}
# Outcome
y1 <- dat_wide$log10VLOAD_2
y2 <- dat_wide$LEU3N_2
y3 <- dat_wide$AGG_PHYS_2
y4 <- dat_wide$AGG_MENT_2


# Design matrix
X1a <- model.matrix(~ hard_drugs_0 + log10VLOAD_0 + age_0 + BMI_0 +
                    smoke_group_0 + educ_group_0 + race_group_0, data = dat_wide)
X1b <- model.matrix(~ hard_drugs_0 + log10VLOAD_0 + age_0 + BMI_0 +
                    smoke_group_0 + educ_group_0 + race_group_0 + adh_group_2,
                   data = dat_wide)
X2a <- model.matrix(~ hard_drugs_0 + LEU3N_0 + age_0 + BMI_0 +
                    smoke_group_0 + educ_group_0 + race_group_0, data = dat_wide)
X2b <- model.matrix(~ hard_drugs_0 + LEU3N_0 + age_0 + BMI_0 +
                    smoke_group_0 + educ_group_0 + race_group_0 + adh_group_2,
                   data = dat_wide)
X3a <- model.matrix(~ hard_drugs_0 + AGG_PHYS_0 + age_0 + BMI_0 +
                    smoke_group_0 + educ_group_0 + race_group_0, data = dat_wide)
X3b <- model.matrix(~ hard_drugs_0 + AGG_PHYS_0 + age_0 + BMI_0 +
                    smoke_group_0 + educ_group_0 + race_group_0 + adh_group_2,
                   data = dat_wide)
X4a <- model.matrix(~ hard_drugs_0 + AGG_MENT_0 + age_0 + BMI_0 +
                    smoke_group_0 + educ_group_0 + race_group_0, data = dat_wide)
X4b <- model.matrix(~ hard_drugs_0 + AGG_MENT_0 + age_0 + BMI_0 +
                    smoke_group_0 + educ_group_0 + race_group_0 + adh_group_2,
                   data = dat_wide)

# Sizes
N1a <- nrow(X1a)
P1a <- ncol(X1a)
N1b <- nrow(X1b)
P1b <- ncol(X1b)
N2a <- nrow(X2a)
P2a <- ncol(X2a)
N2b <- nrow(X2b)
P2b <- ncol(X2b)
N3a <- nrow(X3a)
P3a <- ncol(X3a)
N3b <- nrow(X3b)
P3b <- ncol(X3b)
N4a <- nrow(X4a)
P4a <- ncol(X4a)
N4b <- nrow(X4b)
P4b <- ncol(X4b)

# Priors: Normal(0, 1000) for all betas, Half-normal(0, 1000) for sigma
prior_mean_1a <- rep(0, P1a); prior_sd_1a   <- rep(1000, P1a)
prior_mean_1b <- rep(0, P1b); prior_sd_1b   <- rep(1000, P1b)
prior_mean_2a <- rep(0, P2a); prior_sd_2a   <- rep(1000, P2a)
prior_mean_2b <- rep(0, P2b); prior_sd_2b   <- rep(1000, P2b)
prior_mean_3a <- rep(0, P3a); prior_sd_3a   <- rep(1000, P3a)
prior_mean_3b <- rep(0, P3b); prior_sd_3b   <- rep(1000, P3b)
prior_mean_4a <- rep(0, P4a); prior_sd_4a   <- rep(1000, P4a)
prior_mean_4b <- rep(0, P4b); prior_sd_4b   <- rep(1000, P4b)
sigma_prior_sd <- 1000

# Set up datalists for each model
data_list_1a <- list(
  N = N1a, P = P1a, X = X1a, y = y1,
  prior_mean = prior_mean_1a,
  prior_sd = prior_sd_1a,
  sigma_prior_sd = sigma_prior_sd
)
data_list_1b <- list(
  N = N1b, P = P1b, X = X1b, y = y1,
  prior_mean = prior_mean_1b,
  prior_sd = prior_sd_1b,
  sigma_prior_sd = sigma_prior_sd
)
data_list_2a <- list(
  N = N2a, P = P2a, X = X2a, y = y2,
  prior_mean = prior_mean_2a,
  prior_sd = prior_sd_2a,
  sigma_prior_sd = sigma_prior_sd
)
data_list_2b <- list(
  N = N2b, P = P2b, X = X2b, y = y2,
  prior_mean = prior_mean_2b,
  prior_sd = prior_sd_2b,
  sigma_prior_sd = sigma_prior_sd
)
data_list_3a <- list(
  N = N3a, P = P3a, X = X3a, y = y3,
  prior_mean = prior_mean_3a,
  prior_sd = prior_sd_3a,
  sigma_prior_sd = sigma_prior_sd
)
data_list_3b <- list(
  N = N3b, P = P3b, X = X3b, y = y3,
  prior_mean = prior_mean_3b,
  prior_sd = prior_sd_3b,
  sigma_prior_sd = sigma_prior_sd
)
data_list_4a <- list(
  N = N4a, P = P4a, X = X4a, y = y4,
  prior_mean = prior_mean_4a,
  prior_sd = prior_sd_4a,
  sigma_prior_sd = sigma_prior_sd
)
data_list_4b <- list(
  N = N4b, P = P4b, X = X4b, y = y4,
  prior_mean = prior_mean_4b,
  prior_sd = prior_sd_4b,
  sigma_prior_sd = sigma_prior_sd
)
```
```{r bayesian, echo=FALSE, message=FALSE, results='hide', warning=FALSE, cache=TRUE}
## Bayesian models
fit_vload <- mod$sample(data = data_list_1a, chains = 4, iter_warmup = 2000, iter_sampling = 5000, seed = 123)
fit_cd4 <- mod$sample(data = data_list_2a, chains = 4, iter_warmup = 2000, iter_sampling = 5000, seed = 123)
fit_phys <- mod$sample(data = data_list_3a, chains = 4, iter_warmup = 2000, iter_sampling = 5000, seed = 123)
fit_ment <- mod$sample(data = data_list_4a, chains = 4, iter_warmup = 2000, iter_sampling = 5000, seed = 123)

## Bayesian models with adherence
fit_vload_adh <- mod$sample(data = data_list_1b, chains = 4, iter_warmup = 2000, iter_sampling = 5000, seed = 123)
fit_cd4_adh <- mod$sample(data = data_list_2b, chains = 4, iter_warmup = 2000, iter_sampling = 5000, seed = 123)
fit_phys_adh <- mod$sample(data = data_list_3b, chains = 4, iter_warmup = 2000, iter_sampling = 5000, seed = 123)
fit_ment_adh <- mod$sample(data = data_list_4b, chains = 4, iter_warmup = 2000, iter_sampling = 5000, seed = 123)

## View model outputs
fit_vload
fit_vload_adh
fit_cd4
fit_cd4_adh
fit_phys
fit_phys_adh
fit_ment
fit_ment_adh
```
```{r bayes_tables, echo=FALSE, message=FALSE, results='hide', warning=FALSE, cache=TRUE}
summarize_stan_beta2 <- function(fit, threshold) { # Function to extract Bayesian output
  draws <- fit$draws()
    # Extract full posterior draws
  draws_mat <- as_draws_matrix(draws)
    # Convert draws to matrix form for indexing
  vals <- as.numeric(draws_mat[, "beta[2]"])
    # Pull posterior samples  
  hpd <- hdi(vals, ci = 0.95)
    # Calculate HPDI
  
  tibble(
    Parameter = "beta[2]",
    PM = mean(vals),
    HPDI_2.5 = hpd[1],
    HPDI_97.5 = hpd[2],
    PP = mean(abs(vals) > threshold)
      # Calcualte posterior probability based on clinically meaningful thresholds
  ) # Return results as a tibble
}

model_list <- list(
  `Viral Load (log10 HIV copies/mL blood)` = fit_vload,
  `Viral Load (log10 HIV copies/mL blood)_adh` = fit_vload_adh,
  `CD4+ T Cell Count` = fit_cd4,
  `CD4+ T Cell Count_adh` = fit_cd4_adh,
  `Physical Quality of Life (SF-36 Aggregate)` = fit_phys,
  `Physical Quality of Life (SF-36 Aggregate)_adh` = fit_phys_adh,
  `Mental Quality of Life (SF-36 Aggregate)` = fit_ment,
  `Mental Quality of Life (SF-36 Aggregate)_adh` = fit_ment_adh
) # List of Bayesian models, renamed for tables

thresholds <- c(0.5, 0.5, 50, 50, 2, 2, 2, 2)
  # Clinically meaningful thresholds

tab_bayes <- purrr::map2_df(model_list, thresholds, summarize_stan_beta2, .id = "Model") %>%
    # Apply function to each model
  mutate(
    HPDI = sprintf("(%0.3f, %0.3f)", HPDI_2.5, HPDI_97.5)
  ) %>% # Combine HPDIs and limit to 3 decimal places
  select(Outcome = Model, `Posterior Mean` = PM, `95% HPDI` = HPDI, `Posterior Probability` = PP)
    # Select and rename final columns

tab_bayes_adh <- purrr::map2_df(model_list, thresholds, summarize_stan_beta2, .id = "Model") %>%
    # Apply function to each model
  mutate(
    Outcome = sub("_adh$", "", Model),
    adh     = grepl("_adh$", Model),
      # Select only results from adherence models
    HPDI    = sprintf("(%0.3f, %0.3f)", HPDI_2.5, HPDI_97.5)
      # Combine HPDIs and limit to 3 decimal places
  ) %>%
  select(Outcome, adh, PM, HPDI, PP) %>%
    # Select relevant columns
  pivot_wider(
    names_from  = adh,
    values_from = c(PM, HPDI, PP)
  ) %>% # Pivot to wide and add subscript to note adh 
  rename(
    PM        = PM_FALSE,
    PM_adh    = PM_TRUE,
    HPDI      = HPDI_FALSE,
    HPDI_adh  = HPDI_TRUE,
    PP        = PP_FALSE,
    PP_adh    = PP_TRUE
  ) %>% # Rename columns 
  mutate(
    Percent_Change = 100 * (PM_adh - PM) / abs(PM)
  ) %>% # Calculate impact of adherence
  select(Outcome, `Posterior Mean` = PM_adh, `Change in PM (%)` = Percent_Change, `95% HPDI` = HPDI_adh, `Posterior Probability` = PP_adh)
    # Select and rename final columns
```
```{r comb_tables, echo=FALSE, message=FALSE, results='hide', warning=FALSE, cache=TRUE}
tab_comb <- dplyr::left_join(tab_freq, tab_bayes, by = "Outcome") %>%
    # Combine two tables
  filter(!grepl("_adh$", Outcome)) %>%
    # Filter to models without adherence
  gt() %>%
    # Convert to gt table
  tab_spanner(
    label = "Frequentist",
    columns = c(`Point Estimate`, `95% CI`, `Unadjusted P-Value`)
  ) %>% # Add Frequentist label
  tab_spanner(
    label = "Bayesian",
    columns = c(`Posterior Mean`, `95% HPDI`, `Posterior Probability`)
  ) %>% # Add Bayesian label
  fmt_number(
    columns = c(`Point Estimate`, `Unadjusted P-Value`, `Posterior Mean`, `Posterior Probability`),
    decimals = 3
  ) %>% # Round estimates
  tab_header(
    title = "Comparison of Frequentist and Bayesian Estimates for Associations Between Hard Drug Use and Clinical Outcomes"
  ) %>% 
  tab_footnote(
    footnote = "Comparison of Frequentist and Bayesian regression estimates for the association between baseline hard drug use and year‑2 outcomes. All models adjust for baseline outcome values, age, BMI, smoking status, education, and race. Bayesian models use weakly informative Normal(0, 1000) priors for regression coefficients and half‑Normal(0, 1000) priors for residual SD." 
  ) 

tab_adh <- dplyr::left_join(tab_freq_adh, tab_bayes_adh, by = "Outcome") %>%
    # Combine two tables
  gt() %>%
    # Convert to gt table
  tab_spanner(
    label = "Frequentist",
    columns = c(`Point Estimate`, `Change in PE (%)`, `95% CI`, `Unadjusted P-Value`)
  ) %>% # Add Frequentist label
  tab_spanner(
    label = "Bayesian",
    columns = c(`Posterior Mean`, `Change in PM (%)`, `95% HPDI`, `Posterior Probability`)
  ) %>% # Add Bayesian label
  fmt_number(
    columns = c(`Point Estimate`, `Change in PE (%)`, `Unadjusted P-Value`, `Posterior Mean`, `Change in PM (%)`, `Posterior Probability`),
    decimals = 3
  ) %>% # Round estimates
  tab_header(
    title = "Impact of Adherence Adjustment on Frequentist and Bayesian Estimates for Associations Between Hard Drug Use and Clinical Outcomes"
  ) %>%
  tab_footnote(
    footnote = "Frequentist and Bayesian regression estimates for the association between baseline hard drug use and year‑2 outcomes after adjusting for adherence. All models also adjust for baseline outcome values, age, BMI, smoking status, education, and race. Percent change reflects the relative difference between adherence‑adjusted and non‑adjusted models, calculated for both the frequentist point estimate and the Bayesian posterior mean. Bayesian models use weakly informative Normal(0, 1000) priors for regression coefficients and half‑Normal(0, 1000) priors for residual SD." 
  )
```
```{r fig_raw, echo=FALSE, message=FALSE, results='hide', warning=FALSE, cache=TRUE}
dat_plot <- dat_wide %>%
  select(
    hard_drugs_0,
    log10VLOAD_2,
    LEU3N_2,
    AGG_PHYS_2,
    AGG_MENT_2
  ) %>% # Select outcomes and drugs
  pivot_longer(
    cols = -hard_drugs_0,
    names_to = "Outcome",
    values_to = "Value"
  ) %>% # Pivot to long
  mutate(
    hard_drugs_0 = factor(hard_drugs_0, labels = c("No Hard Drugs", "Hard Drugs")),
    Outcome = recode(
      Outcome,
      log10VLOAD_2 = "Viral Load (log10 HIV copies/mL blood)",
      LEU3N_2      = "CD4+ T Cell Count",
      AGG_PHYS_2   = "Physical Quality of Life (SF-36 Aggregate)",
      AGG_MENT_2   = "Mental Quality of Life (SF-36 Aggregate)"
    )
  ) # Rename for figure

fig1 <- ggplot(dat_plot, aes(x = hard_drugs_0, y = Value)) +
    # Plot outcomes by baseline hard drug use
  geom_violin(fill = "gray85", color = "gray60") +
    # Raw data as violin
  geom_jitter(alpha = 0.25, width = 0.15, color = "gray40") +
    # Violin transparent
  facet_wrap(~ Outcome, scales = "free_y") +
    # Facet by outcome
  labs(
    x = "Hard Drug Use at Baseline",
    y = "Outcome Value",
    title = "Observed 2nd Year Outcomes by Baseline Hard Drug Use"
  ) +
  theme_bw() +
  theme(
    strip.text = element_text(size = 11),
    legend.position = "none"
  )
```
```{r fig_models, echo=FALSE, message=FALSE, results='hide', warning=FALSE, cache=TRUE}
forest_df <-
  purrr::map_df(mod_list, summarize_lm_beta2, .id = "Model") %>%
    # Frequentist
  mutate(
    Adjustment = ifelse(grepl("_adh$", Model), "Adherence-adjusted", "Unadjusted"),
    Outcome = sub("_adh$", "", Model),
    Method = "Frequentist"
  ) %>% # Add labels 
  transmute(
    Outcome,
    Method,
    Adjustment,
    Estimate = PE,
    lower = CI_2.5,
    upper = CI_97.5
  ) %>% # Rename 
  
  bind_rows(
    purrr::map2_df(model_list, thresholds, summarize_stan_beta2, .id = "Model") %>%
        # Bayesian
      mutate(
        Adjustment = ifelse(grepl("_adh$", Model), "Adherence-adjusted", "Unadjusted"),
        Outcome = sub("_adh$", "", Model),
        Method = "Bayesian"
      ) %>% # Add labels
      transmute(
        Outcome,
        Method,
        Adjustment,
        Estimate = PM,
        lower = HPDI_2.5,
        upper = HPDI_97.5
      ) # Rename
  ) %>% # Add Bayesian outcomes
  
  mutate(
    Outcome = recode(
      Outcome,
      log10VLOAD_2 = "Viral Load (log10 HIV copies/mL blood)",
      LEU3N_2      = "CD4+ T Cell Count",
      AGG_PHYS_2   = "Physical Quality of Life (SF-36 Aggregate)",
      AGG_MENT_2   = "Mental Quality of Life (SF-36 Aggregate)"
    ), # Rename outcomes
    cm_diff = case_when(
      Outcome == "Viral Load (log10 HIV copies/mL blood)" ~ 0.5,
      Outcome == "CD4+ T Cell Count" ~ 50,
      Outcome == "Physical Quality of Life (SF-36 Aggregate)" ~ 2,
      Outcome == "Mental Quality of Life (SF-36 Aggregate)" ~ 2
    ), # Add clinically meaningful differences
    Outcome = factor(
      Outcome,
      levels = c(
        "Viral Load (log10 HIV copies/mL blood)",
        "CD4+ T Cell Count",
        "Physical Quality of Life (SF-36 Aggregate)",
        "Mental Quality of Life (SF-36 Aggregate)"
      ) # Order facets
    )
  )

fig2 <- ggplot(forest_df, aes(x = Estimate, y = Method, color = Method, shape = Adjustment)) +
    # Forest plot of outcomes
  geom_rect(
    aes(
      xmin = -cm_diff, xmax = cm_diff,
      ymin = -Inf, ymax = Inf
    ),
    fill = "grey90", alpha = 0.3,
    inherit.aes = FALSE
  ) + # Add shaded area for clinical differences
  geom_vline(xintercept = 0, linetype = "dashed") +
    # Add line at 0
  geom_point(size = 3, position = position_dodge(width = 0.5)) +
    # Add outcomes
  geom_errorbar(
    aes(xmin = lower, xmax = upper),
    width = 0.2,
    position = position_dodge(width = 0.5)
  ) + # Add intervals
  facet_wrap(~ Outcome, scales = "free_x") +
    # Facet by outcome
  scale_shape_manual(values = c(
    "Unadjusted" = 16,
    "Adherence-adjusted" = 17
  )) + # Add shapes based on adherence 
  labs(
    x = "Adjusted Effect of Baseline Hard Drug Use",
    y = "",
    title = "Adjusted Associations Between Baseline Hard Drug Use and 2nd Year Outcomes",
    color = "Model Type",
    shape = "Adjustment",
  ) +
  theme_bw() +
  theme(
    legend.position = "bottom",
    strip.text = element_text(size = 11),
  )
```

## Introduction

This project uses data from the Multicenter AIDS Cohort Study, an ongoing prospective cohort that studies HIV-1 infection among homosexual and bisexual men in four major US cities. This report analyzed a subset of 463 MACS participants with viral load (log10 HIV copies/mL blood), CD4+ T cell count, and SF-36 aggregate physical and mental quality‑of‑life measured at baseline (pre‑HAART initiation) and two years after starting HAART. Baseline information on hard drug use, defined as injection drugs or illicit heroin/opiate use, was available along with and other relevant covariates including age, BMI, smoking status, education, and race/ethnicity. 

The primary question of interest was whether individuals who reported hard drug use at baseline experienced different treatment outcomes two years after starting HAART compared to those who did not after adjusting for baseline covariates. The outcomes were viral load (log10 HIV copies/mL blood), CD4+ T cell count, and SF-36 aggregate physical and mental quality‑of‑life. The covariates were age, BMI, smoking status, education, and race/ethnicity. The secondary analysis will descriptively assess whether the difference or lack of difference in treatment response between hard drug users and non hard drug users after adjusting for baseline covariates can be explained by year-2 HAART adherence. These two questions were both assessed from a frequentist and a Bayesian framework.

## Methods

All analyses were conducted in R v4.5.2 (2025-10-31 ucrt). BMI values coded as -1 or 999 and two implausible values over 250 were set to missing. Categorical covariates were collapsed as follows to ensure adequate sample sizes by category: smoking status (current vs former/non smoker), education status (no college degree vs college degree and above), race/ethnicity (white non-Hispanic vs other), and adherence (95% and above vs <95%). Viral load was converted to the log10 scale. Participants were removed if they were missing baseline  after removing any individuals with incomplete baseline viral load, CD4+ T cell count, physical and mental quality of life scores, age, BMI, smoking status, education, or race/ethnicity. Participants were also removed if they were missing year two outcomes or year two adherence. The final sample size was 463 individuals.

For each outcome at year 2, linear regression was used to identify the association with hard drug use at baseline, adjusting for age, BMI, smoking status, education, and race/ethnicity. Frequentist models reported point estimates, 95\% confidence intervals (CI), and unadjusted p-values. Bayesian models were implemented using STAN with vaguely informative priors as prior knowledge is limited: N(0,1000) for regression coefficients and half-normal(0,1000) for residual SD. The intercept was not centered due to the large variance and model fit was diagnosed with R-hat and ESS. Bayesian models reported posterior means, 95% highest posterior density intervals (HPDI), and posterior probabilities. Posterior probabilities were calculated from clinically meaningful differences for each outcome (two-sided), defined as follows: 0.5 log10 (viral load), 50 cells (CD4+ T cell count), and 2 points (quality of life). These models were additionally fit including year 2 adherence and descriptively compared to the original models. 

## Results



## Appendix

Code Directory: https://github.com/eweible/BIOS6624.git

File name: P1/Code/P1_Report.Rmd

Data Directory: C:/Users/weiblee/OneDrive/Documents/Anschutz/Spring 2026/6624_Advanced/Weible_Advanced/P1/DataRaw
