---
title: "P3_Data_Exploration"
output: html_document
date: "2026-04-01"
editor_options: 
  chunk_output_type: inline
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
library(dplyr)
library(tidyr)
library(survival)
library(survminer)
library(here)
library(table1)
library(ggplot2)
library(gt)
dat_raw <- read.csv(here::here("P3/DataRaw", "frmgham2.csv"))
```

```{r}
View(dat_raw)
```

# Data cleaning
```{r}
dat_10 <- dat_raw %>%
  filter(!(STROKE == 1 & TIMESTRK == 0)) %>%
    # Remove baseline stroke
  mutate(
    time_yrs = TIMESTRK / 365.25,
    time_10 = pmin(time_yrs, 10),
    stroke_event = if_else(STROKE == 1 & time_yrs <= 10, 1, 0)
  ) # Adjust time vars
dat_10 <- dat_10 %>%
  dplyr::select(
    RANDID, STROKE, SEX, time_10, stroke_event, AGE, DIABETES, SYSBP, PREVCHD, 
    BPMEDS, CURSMOKE, TOTCHOL, BMI
  ) # Restrict to needed vars
dat_10 <- dat_10 %>%
  mutate(
    SEX = factor(SEX, labels = c("Male", "Female")),
    DIABETES = factor(DIABETES, levels = c(0,1), labels = c("No","Yes")),
    CURSMOKE = factor(CURSMOKE, levels = c(0,1), labels = c("No","Yes")),
    PREVCHD = factor(PREVCHD, levels = c(0,1), labels = c("No","Yes")),
    BPMEDS = factor(BPMEDS, levels = c(0,1), labels = c("No","Yes"))
  ) # Convert to factors
dat_cc <- dat_10 %>%
  filter(!is.na(time_10), !is.na(stroke_event)) %>%
  filter(complete.cases(
    AGE, SYSBP, DIABETES, PREVCHD, BPMEDS, CURSMOKE, TOTCHOL, BMI
  )) # Complete case dataset for all vars
```

# Kaplan-Meier curves and log-rank tests
```{r}
surv_object <- Surv(time = dat_cc$time_10, event = dat_cc$stroke_event)

## By sex
fit_sex <- survfit(surv_object ~ SEX, data = dat_cc)
ggsurvplot(fit_sex, data = dat_cc, ylim = c(0.90, 1.00),
           xlab = "Years", ylab = "Stroke-free survival")$plot
survdiff(surv_object ~ SEX, data = dat_cc)

## By diabetes
fit_diab <- survfit(surv_object ~ DIABETES, data = dat_cc)
ggsurvplot(fit_diab, data = dat_cc, ylim = c(0.90, 1.00),
           xlab = "Years", ylab = "Stroke-free survival")$plot
survdiff(surv_object ~ DIABETES, data = dat_cc)

## By high BP
dat_cc <- dat_cc %>%
  mutate(HIGHBP = if_else(SYSBP >= 160, 1, 0))
fit_bp <- survfit(Surv(time_10, stroke_event) ~ HIGHBP, data = dat_cc)
ggsurvplot(fit_bp, data = dat_cc, ylim = c(0.90, 1.00),
           xlab = "Years", ylab = "Stroke-free survival")$plot
survdiff(Surv(time_10, stroke_event) ~ HIGHBP, data = dat_cc)
```
**Summary:**
* Sex: KM curves overlap throughout (no difference in survival, supported by log-rank X^2 = 0.4, p = 0.5).
* Diabetes: KM curves overlap mostly in the first 5 years (small numbers of early events). Diverge strongly later (log-rank: X^2 = 34.3, p < 0.001).
* BP: Strong visual difference on KM curves and log-rank (high BP has much higher stroke rate, separate cleanly). X^2 = 149, p << 0.001.

# Table 1:
```{r}
table1(~ AGE + SYSBP + DIABETES + PREVCHD + BPMEDS + CURSMOKE + TOTCHOL + BMI 
       | SEX, data = dat_cc)
```

# Cox models and PH checks
```{r}
## Males
dat_male <- dat_cc %>% filter(SEX == "Male")
dat_female <- dat_cc %>% filter(SEX == "Female")

cox_male_full <- coxph(Surv(time_10, stroke_event) ~ AGE + SYSBP + DIABETES + 
                         PREVCHD + BPMEDS + CURSMOKE + TOTCHOL + BMI, 
                       data = dat_male) # Full model
cox_male_min <- coxph(Surv(time_10, stroke_event) ~ AGE + DIABETES + SYSBP,
                      data = dat_male) # Smallest model
cox_male_step <- step(cox_male_full, direction = "backward", 
                      scope = list(lower = formula(cox_male_min),
                                   upper = formula(cox_male_full)), k = 2)
  # Backward selection (AIC)
    # Male: Age, SysBP, Diabetes, BP Meds, Cur Smoke, Tot Chol
summary(cox_male_step)

ph_male <- cox.zph(cox_male_step)
ph_male
ggcoxzph(ph_male)
  # Check PH

## Females
cox_female_full <- coxph(Surv(time_10, stroke_event) ~ AGE + SYSBP + DIABETES +
                           PREVCHD + BPMEDS + CURSMOKE + TOTCHOL + BMI,
                         data = dat_female) # Full model
cox_female_min <- coxph(Surv(time_10, stroke_event) ~ AGE + DIABETES + SYSBP,
                      data = dat_female) # Smallest model
cox_female_step <- step(cox_female_full, direction = "backward", , 
                      scope = list(lower = formula(cox_female_min),
                                   upper = formula(cox_female_full)), k = 2)
  # Backward selection (AIC)
    # Female: Age, SysBP, Diabetes, BP Meds, Prev CHD
summary(cox_female_step)

ph_female <- cox.zph(cox_female_step)
ph_female
ggcoxzph(ph_female)
  # Check PH
```

**Summary:**
* Males: diabetes and sysBP dominate the stroke risk, age and BP meds also contribute, CURSMOKE and TOTCHOL are marginal/borderline may not actually need to be included
* Females: diabetes is a weaker predictor than for men, sysBP still dominates. Age, BP meds, and PREVCHD also contribute
* PH tests look good overall. Only diabetes mildly violates for men (p = 0.008, more curvy), globally is good for each.

# Predicted 10 year probabilities
```{r}
ages <- c(40, 50, 60)

## Males
profiles_male <- expand.grid(
  AGE = ages,
  SYSBP = c(120, 160),
    # Normal vs. high
  DIABETES = c("No", "Yes")
) %>% 
  mutate(
    PREVCHD = "No",
    BPMEDS = "No",
    CURSMOKE = "No",
    TOTCHOL = mean(dat_male$TOTCHOL, na.rm = TRUE),
    BMI = mean(dat_male$BMI, na.rm = TRUE)
  ) # Profiles for the combos of BP/diabetes
profiles_male_other <- expand.grid(
  AGE = ages
) %>%
  mutate(
    SYSBP = 120,
    DIABETES = "No",
    PREVCHD = "No",
    BPMEDS = "Yes",
    CURSMOKE = "No",
    TOTCHOL = mean(dat_male$TOTCHOL, na.rm = TRUE),
    BMI = mean(dat_male$BMI, na.rm = TRUE)
  ) # Other profile: BP meds
profiles_male <- bind_rows(profiles_male, profiles_male_other)
  # Combine profiles
base_male <- survfit(cox_male_step, newdata = profiles_male)
  # Baseline survival
pred_male <- summary(base_male, times = 10)
  # Extract survival at 10 years
stroke_prob_10_male <- 1 - pred_male$surv
  # 10-year stroke probability = 1 - S(10)

## Females
profiles_female <- expand.grid(
  AGE = ages,
  SYSBP = c(120, 160),
    # Normal vs. high
  DIABETES = c("No", "Yes")
) %>% 
  mutate(
    PREVCHD = "No",
    BPMEDS = "No",
    CURSMOKE = "No",
    TOTCHOL = mean(dat_female$TOTCHOL, na.rm = TRUE),
    BMI = mean(dat_female$BMI, na.rm = TRUE)
  ) # Profiles for the combos of BP/diabetes
profiles_female_other <- expand.grid(
  AGE = ages
) %>%
  mutate(
    SYSBP = 120,
    DIABETES = "No",
    PREVCHD = "No",
    BPMEDS = "Yes",
    CURSMOKE = "No",
    TOTCHOL = mean(dat_female$TOTCHOL, na.rm = TRUE),
    BMI = mean(dat_female$BMI, na.rm = TRUE)
  ) # Other profile: BP meds
profiles_female <- bind_rows(profiles_female, profiles_female_other)
  # Combine profiles
base_female <- survfit(cox_female_step, newdata = profiles_female)
  # Baseline survival
pred_female <- summary(base_female, times = 10)
  # Extract survival at 10 years
stroke_prob_10_female <- 1 - pred_female$surv
  # 10-year stroke probability = 1 - S(10)
```

```{r}
male_tbl <- profiles_male %>%
  mutate(
    prob = as.numeric(stroke_prob_10_male) * 100
  ) %>% # Convert to percent
  mutate(
    Profile = case_when(
      BPMEDS == "Yes" ~ "Other",
      SYSBP == 120 & DIABETES == "No" ~ "Baseline",
      SYSBP == 160 & DIABETES == "No" ~ "High BP",
      SYSBP == 120 & DIABETES == "Yes" ~ "Diabetes",
      SYSBP == 160 & DIABETES == "Yes" ~ "Both"
    )
  ) %>%
  select(AGE, Profile, prob)
male_table_final <- male_tbl %>%
  pivot_wider(names_from = Profile, values_from = prob) %>%
  arrange(AGE)
female_tbl <- profiles_female %>%
  mutate(
    prob = as.numeric(stroke_prob_10_female) * 100
  ) %>% # Convert to percent
  mutate(
    Profile = case_when(
      BPMEDS == "Yes" ~ "Other",
      SYSBP == 120 & DIABETES == "No"  ~ "Baseline",
      SYSBP == 160 & DIABETES == "No"  ~ "High BP",
      SYSBP == 120 & DIABETES == "Yes" ~ "Diabetes",
      SYSBP == 160 & DIABETES == "Yes" ~ "Both"
    )
  ) %>%
  select(AGE, Profile, prob)
female_table_final <- female_tbl %>%
  pivot_wider(names_from = Profile, values_from = prob) %>%
  arrange(AGE)
male_table_finaL <- male_table_final %>% mutate(SEX = "Male")
female_table_final <- female_table_final %>% mutate(SEX = "Female")

combined_table <- bind_rows(female_table_final, male_table_final) %>%
  relocate(SEX) %>%
  arrange(SEX, AGE) %>%
  gt(rowname_col = "AGE", groupname_col = "SEX") %>%
  tab_header(
    title = "10-year Stroke Risk by Sex and Clinical Profile"
  ) %>%
  fmt_number(
    columns = where(is.numeric),
    decimals = 1
  ) %>%
  cols_label(
    AGE = "Age",
    Baseline = "Baseline",
    `High BP` = "High BP",
    Diabetes = "Diabetes",
    Both = "Both",
    Other = "Other"
  ) %>%
  tab_style(
    style = cell_text(weight = "bold"),
    locations = cells_column_labels(everything())
  ) %>%
  tab_options(
    row_group.as_column = FALSE
  )



male_heat <- profiles_male %>%
  mutate(
    SEX = "Male",
    prob = as.numeric(stroke_prob_10_male) * 100,
    SYSBP = factor(SYSBP),
    DIABETES = factor(DIABETES),
    BPMEDS = factor(BPMEDS)
  )

female_heat <- profiles_female %>%
  mutate(
    SEX = "Female",
    prob = as.numeric(stroke_prob_10_female) * 100,
    SYSBP = factor(SYSBP),
    DIABETES = factor(DIABETES),
    BPMEDS = factor(BPMEDS)
  )

heat_all <- bind_rows(male_heat, female_heat)
ggplot(heat_all, aes(x = SYSBP, y = AGE, fill = prob)) +
  geom_tile(color = "white") +
  facet_grid(SEX ~ DIABETES, labeller = label_both) +
  scale_fill_gradient(low = "lightblue", high = "darkred",
                      name = "10-year risk (%)") +
  labs(
    title = "10-year Stroke Risk by Sex",
    x = "Systolic BP",
    y = "Age"
  ) +
  theme_minimal()
```

# Time-varying patterns (BP and diabetes)
```{r}
dat_long <- dat_raw %>%
  filter(PERIOD %in% c(1,2,3)) %>%
  select(RANDID, PERIOD, SEX, SYSBP, DIABETES)

dat_long %>%
  group_by(SEX, PERIOD) %>%
  summarise(mean_SYSBP = mean(SYSBP, na.rm = TRUE),
            sd_SYSBP   = sd(SYSBP, na.rm = TRUE))

dat_long %>%
  group_by(SEX, PERIOD) %>%
  summarise(diabetes_prev = mean(DIABETES == 1, na.rm = TRUE))

ggplot(dat_long, aes(x = PERIOD, y = SYSBP, color = SEX)) +
  stat_summary(fun = mean, geom = "line") +
  labs(x = "Exam Period", y = "Mean Systolic BP")

ggplot(dat_long, aes(x = PERIOD, y = DIABETES, color = SEX)) +
  stat_summary(fun = mean, geom = "line") +
  labs(x = "Exam Period", y = "Diabetes Prevalence")
```


