---
title: "P0_Draft_Analysis"
output: html_document
date: "2026-01-29"
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
library(dplyr)
library(tidyr)
library(lme4)
library(ggplot2)
raw_dat <- read.csv(here::here("P0/DataRaw", "Project0_Clean_v2.csv"))
```

# Data Manipulation

```{r}
analysis_dat <- raw_dat %>%
  select(
    -Booklet..Sample.interval,
    -Booklet..Sample.interval.Decimal.Time..mins.,
    -MEMs..Sample.interval,
    -MEMs..Sample.interval.Decimal.Time..mins.,
    -Cortisol..ug.dl.,
    -DHEA..pg.dl.
  ) %>% # Remove unwanted columns
  
  group_by(SubjectID) %>%
  mutate(
    dhea_upper_hits = sum(DHEA..nmol.L. == 5.205, na.rm = TRUE)
  ) %>%
  ungroup() %>% # Identify number of DHEA values at the detection limit (5.205)
  { 
    removed_info <<- distinct(.[.$dhea_upper_hits > 2, c("SubjectID", "dhea_upper_hits")])
    . # Return to data
  } %>% # Save IDs & number of hits of participants to be removed
  filter(!SubjectID %in% participants_to_remove) %>%
    # Remove participants with >2 detection limit hits across all their rows
  
  mutate(
    Cortisol..nmol.L. = ifelse(Cortisol..nmol.L. > 80, NA, Cortisol..nmol.L.),
      # Remove cortisol > 80 nmol/L
    DHEA..nmol.L. = ifelse(DHEA..nmol.L. == 5.205, NA, DHEA..nmol.L.)
      # Remove DHEA at the detection limit
  ) %>% 
  select(-dhea_upper_hits)
    # Drop helper column
analysis_dat <- analysis_dat %>%
  mutate(
    Booket..Clock.Time = na_if(trimws(Booket..Clock.Time), ""),
    MEMs..Clock.Time = na_if(trimws(MEMs..Clock.Time), ""),
    Sleep.Diary.reported.wake.time = na_if(trimws(Sleep.Diary.reported.wake.time), "")
  ) %>% # Convert empty strings to NAs and trim leading spaces
  mutate(
    booklet_time = as.POSIXct(
      paste(Collection.Date, Booket..Clock.Time),
      format = "%m/%d/%Y %H:%M"
        # MM/DD/YYYY format
    ),
    cap_time = as.POSIXct(
      paste(Collection.Date, MEMs..Clock.Time),
      format = "%m/%d/%Y %H:%M"
        # MM/DD/YYYY format
    ),
    reported_wake_time = as.POSIXct(
      paste(Collection.Date, Sleep.Diary.reported.wake.time),
      format = "%m/%d/%Y %H:%M"
        # MM/DD/YYYY format
    )
  ) %>% # Combine date + time into full timestamps
  group_by(SubjectID, Collection.Date) %>%   
    # Group by subject/day
  fill(reported_wake_time, .direction = "down") %>%  
    # Fill missing wake times so no NAs
  ungroup() %>%
  mutate(
    wake_diff_booklet = as.numeric(booklet_time - reported_wake_time, units = "mins"),
      # Difference between booklet time and wake time
    wake_diff_cap = as.numeric(cap_time - reported_wake_time, units = "mins"),
      # Difference between cap time and wake time
    diff_booklet_cap = as.numeric(booklet_time - cap_time, units = "mins")
      # Difference between booklet time and cap time
  ) # Create new variables as times in mins
```

# RQ1

What is the agreement between the subject’s recordings of sampling times compared to the times recorded by an electronic monitoring cap?

Answer this question with:

- LMM 
  - Outcome is book min from wake
  - Predictor is cap time from wake
  - Include random intercept for subject
  - If perfect agreement, intercept will be 0 and slope will be 1

```{r}
rq1_agreement <- lmer(wake_diff_booklet ~ wake_diff_cap + (1 | SubjectID), data = analysis_dat)
  # LMM
summary(rq1_agreement)
```
Visually, the intercept and slope are not equal to 0 and 1. Also, the t-values reveal that neither is statistically different from 0 (0.888 and 0.796). I would still like to verify that the slope is not equal to 1 statistically. 

```{r}
car::linearHypothesis(rq1_agreement, "wake_diff_cap = 1")
```

There is no evidence that the intercept differs significantly from 0. Thus, there is no evidence of systematic bias such as one measure being always earlier or always later than the other. However, we can reject the null hypothesis that the slope is not different from 1 and conclude that the relationship between cap and booklet time from wake does not resemble the 1:1 line the investigator described in the desired scatterplot. Rather, changes in cap time do not translate to changes in booklet time as there is no evidence that the slope differs significantly from 0.

# TODO - update interpretation based on updated data manipulation

# RQ2: 

Are subjects adhering accurately to the +30min and +10 hour sampling times required by a study protocol?

Answer this question with:
- Descriptive statistics
  - What proportion falls within +/- 7.5min and what proportion falls within +/- 15 min? 
  
```{r}
analysis_dat %>%
  summarise(
    n = sum(!is.na(diff_booklet_cap)),
    n_missing = sum(is.na(diff_booklet_cap)),
    perc_within_7.5  = mean(abs(diff_booklet_cap) <= 7.5, na.rm = TRUE) * 100,
    perc_within_15   = mean(abs(diff_booklet_cap) <= 15,  na.rm = TRUE) * 100
  ) #Summarize adherence overall 

analysis_dat %>%
  filter(Collection.Sample %in% c(2, 4)) %>%
  group_by(Collection.Sample) %>%
  summarise(
    n = sum(!is.na(diff_booklet_cap)),
    n_missing = sum(is.na(diff_booklet_cap)),
    perc_within_7.5  = mean(abs(diff_booklet_cap) <= 7.5, na.rm = TRUE) * 100,
    perc_within_15   = mean(abs(diff_booklet_cap) <= 15,  na.rm = TRUE) * 100
  ) # Summarize adherence only for 30min and 10hr samples
```

# RQ3:

What are the changes of cortisol and DHEA over time?

Answer this question with:
- Piecewise LMM with knot at 30 min
  - Use booklet time because fewer missing values and matches with sleep diary
  - Outcome is cortisol or DHEA
  - Primary predictor is book time from wake
  
```{r}
# Split time into piecewise slopes 
analysis_dat$time_pre30  <- pmin(analysis_dat$wake_diff_booklet, 30)
analysis_dat$time_post30 <- pmax(analysis_dat$wake_diff_booklet - 30, 0)

rq3_cortisol <- lmer(Cortisol..nmol.L. ~ time_pre30 + time_post30 + (1 | SubjectID), data = analysis_dat)
summary(rq3_cortisol)
```

Based on this model, the estimated cortisol level at wake time (0 min) is 7.222 nmol/L (intercept). The slope from 0-30min is 0.025 nmol/L/min, but it is not strongly significant (t = 1.094). This us unexpected because we were expecting to see a sharp rise in the first 30 min. Then, the slope after 30min is -0.009 nmol/L/min with strong statistical significance (t = -8.101). This can be visualized in the plot below. 
  
```{r}
cort_dat <- rq3_cortisol@frame %>%
  mutate(row_id = as.numeric(rownames(rq3_cortisol@frame)))  
  # Extract data used in model
cort_dat$pred_cortisol <- predict(rq3_cortisol, re.form = NA)
  # Add predictions (population level)

analysis_dat <- analysis_dat %>%
  mutate(row_id = row_number())
plot_dat <- cort_dat %>%
  left_join(
    analysis_dat %>% select(row_id, wake_diff_booklet),
    by = "row_id"
  ) # Merge time variable from full dataset

ggplot(plot_dat, aes(x = wake_diff_booklet, y = Cortisol..nmol.L.)) +
  geom_point(alpha = 0.4) +
  geom_line(aes(y = pred_cortisol), color = "blue", linewidth = 1) +
  geom_vline(xintercept = 30, linetype = "dashed", color = "red") +
  labs(
    x = "Booklet Time Since Wake (min)",
    y = "Cortisol (nmol/L)",
    title = "Predicted Cortisol (Knot at 30 min)"
  ) +
  theme_minimal()
```


