---
title: "P0_Data_Exploration"
output: html_document
date: "2026-01-26"
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
library(dplyr)
library(naniar)
raw_dat <- read.csv(here::here("P0/DataRaw", "Project0_Clean_v2.csv"))
```

# Glimpsing Data

Initial overview:
```{r}
View(raw_dat)
summary(raw_dat)
glimpse(raw_dat)
```
There does appear to be some missing data, either coded as an empty string or as NA. Collection.Date, Collection.Sample, times, and measurements appear to be in the correct format. Unsure about the MEMs, would have to explore further. Perhaps ask clarifying question about how they were determined - by hand for MEMs Sample interval but automatically for the decimal time version? 

Missing data:
```{r}
naniar::gg_miss_var(raw_dat)
naniar::vis_miss(raw_dat)
```

Appears as though the majority of the missingness is from MEMs Sample interval decimal time (over 175 missing). Is there a reason for this??
I wonder if some of it is due to the booklet sample interval decimal time, but there are only a few over 50 missing there, so that cannot be the only reason.

Why are there some missing responses for cortisol and DHEA? Appears as though they are for the same patients, did they miss a visit or collection time? 

Verifying correct values:
```{r}
table(raw_dat$Collection.Sample)
table(raw_dat$DAYNUMB)

hist(raw_dat$MEMs..Sample.interval.Decimal.Time..mins.)

hist(raw_dat$Cortisol..nmol.L.)
hist(raw_dat$DHEA..nmol.L.)
hist(raw_dat$Cortisol..ug.dl.)
hist(raw_dat$DHEA..pg.dl.)
```
What is a reasonable MEMs sample interval decimal time (mins)? There appears to be a wide variety (~20-920)
What is a reasonable level for cortisol and DHEA for each unit? There appears to be more variety for DHEA than cortisol, is this expected?


# Drafting Data Management/Descriptive Stats

TODO: 
- Remove sample interval variables & units for DHEA and Cortisol that are not nmol/L
- Look into rates of missing data
- Create new variable giving the difference between cap and booklet time
- Create scatterplot showing booklet vs. cap time since waking
- Flag levels of cortisol that are over 80, max DHEA value. Exclude individuals with multiple DHEA that are near detection limit
- Plot cortisol and DHEA levels with bar at 30 min

Data Manipulation:
```{r}
analysis_dat <- raw_dat %>%
  select(
    -Booklet..Sample.interval,
    -Booklet..Sample.interval.Decimal.Time..mins.,
    -MEMs..Sample.interval,
    -MEMs..Sample.interval.Decimal.Time..mins.,
    -Cortisol..ug.dl.,
    -DHEA..pg.dl.
  ) %>% # Remove unwanted columns
  
  group_by(SubjectID) %>%
  mutate(
    dhea_upper_hits = sum(DHEA..nmol.L. == 5.205, na.rm = TRUE)
  ) %>%
  ungroup() %>% # Identify number of DHEA values at the detection limit (5.205)
  { 
    removed_info <<- distinct(.[.$dhea_upper_hits > 2, c("SubjectID", "dhea_upper_hits")])
    . # Return to data
  } %>% # Save IDs & number of hits of participants to be removed
  filter(!SubjectID %in% participants_to_remove) %>%
    # Remove participants with >2 detection limit hits across all their rows
  
  mutate(
    Cortisol..nmol.L. = ifelse(Cortisol..nmol.L. > 80, NA, Cortisol..nmol.L.),
      # Remove cortisol > 80 nmol/L
    DHEA..nmol.L. = ifelse(DHEA..nmol.L. == 5.205, NA, DHEA..nmol.L.)
      # Remove DHEA at the detection limit
  ) %>% 
  select(-dhea_upper_hits)
    # Drop helper column
  
# Print about removal to share with investigator
cat("Number of participants removed:", nrow(removed_info), "\n")
cat("Removed participant ID(s):", removed_info$SubjectID, "\n")
cat("Number of DHEA values at detection limit:", removed_info$dhea_upper_hits)
```
Adjusting time variables:
```{r}
analysis_dat <- analysis_dat %>%
  mutate(
    Booket..Clock.Time = na_if(trimws(Booket..Clock.Time), ""),
    MEMs..Clock.Time = na_if(trimws(MEMs..Clock.Time), ""),
    Sleep.Diary.reported.wake.time = na_if(trimws(Sleep.Diary.reported.wake.time), "")
  ) %>% # Convert empty strings to NAs and trim leading spaces
  mutate(
    booklet_time = as.POSIXct(
      paste(Collection.Date, Booket..Clock.Time),
      format = "%m/%d/%Y %H:%M"
        # MM/DD/YYYY format
    ),
    cap_time = as.POSIXct(
      paste(Collection.Date, MEMs..Clock.Time),
      format = "%m/%d/%Y %H:%M"
        # MM/DD/YYYY format
    ),
    reported_wake_time = as.POSIXct(
      paste(Collection.Date, Sleep.Diary.reported.wake.time),
      format = "%m/%d/%Y %H:%M"
        # MM/DD/YYYY format
    )
  ) %>% # Combine date + time into full timestamps
  mutate(
    wake_diff_booklet = as.numeric(booklet_time - reported_wake_time, units = "mins"),
      # Difference between booklet time and wake time
    wake_diff_cap = as.numeric(cap_time - reported_wake_time, units = "mins"),
      # Difference between cap time and wake time
    diff_booklet_cap = as.numeric(booklet_time - cap_time, units = "mins")
      # Difference between booklet time and cap time
  ) # Create new variables as times in mins
```

