---
output:
  bookdown::pdf_document2:
    toc: false
    number_sections: false
documentclass: article
geometry: margin=1in
fontsize: 12pt
linestretch: 2
fig.cap: true
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
library(dplyr)
library(tidyr)
library(lmerTest)
library(ggplot2)
library(ggforce)
library(gtsummary)
raw_dat <- read.csv(here::here("P0/DataRaw", "Project0_Clean_v2.csv"))
  # Read in dataset 
```

```{r data_manip, include=FALSE}
analysis_dat <- raw_dat %>%
  select(
    -Booklet..Sample.interval,
    -Booklet..Sample.interval.Decimal.Time..mins.,
    -MEMs..Sample.interval,
    -MEMs..Sample.interval.Decimal.Time..mins.,
    -Cortisol..ug.dl.,
    -DHEA..pg.dl.
  ) %>% # Remove unwanted columns
  
  group_by(SubjectID) %>%
  mutate(
    dhea_upper_hits = sum(DHEA..nmol.L. == 5.205, na.rm = TRUE)
  ) %>%
  ungroup() %>% # Identify number of DHEA values at the detection limit (5.205)
  { 
    removed_info <<- distinct(.[.$dhea_upper_hits > 2, c("SubjectID", "dhea_upper_hits")])
    . # Return to data
  } %>% # Save IDs & number of hits of participants to be removed
  filter(!SubjectID %in% removed_info) %>%
    # Remove participants with >2 detection limit hits across all their rows
  
  mutate(
    Cortisol..nmol.L. = ifelse(Cortisol..nmol.L. > 80, NA, Cortisol..nmol.L.),
      # Remove cortisol > 80 nmol/L
    DHEA..nmol.L. = ifelse(DHEA..nmol.L. == 5.205, NA, DHEA..nmol.L.)
      # Remove DHEA at the detection limit
  ) %>% 
  select(-dhea_upper_hits)
    # Drop helper column

# # Print about removal to share with investigator
# cat("Number of participants removed:", nrow(removed_info), "\n")
# cat("Removed participant ID(s):", removed_info$SubjectID, "\n")
# cat("Number of DHEA values at detection limit:", removed_info$dhea_upper_hits)

analysis_dat <- analysis_dat %>%
  mutate(
    Booket..Clock.Time = na_if(trimws(Booket..Clock.Time), ""),
    MEMs..Clock.Time = na_if(trimws(MEMs..Clock.Time), ""),
    Sleep.Diary.reported.wake.time = na_if(trimws(Sleep.Diary.reported.wake.time), "")
  ) %>% # Convert empty strings to NAs and trim leading spaces
  mutate(
    booklet_time = as.POSIXct(
      paste(Collection.Date, Booket..Clock.Time),
      format = "%m/%d/%Y %H:%M"
        # MM/DD/YYYY format
    ),
    cap_time = as.POSIXct(
      paste(Collection.Date, MEMs..Clock.Time),
      format = "%m/%d/%Y %H:%M"
        # MM/DD/YYYY format
    ),
    reported_wake_time = as.POSIXct(
      paste(Collection.Date, Sleep.Diary.reported.wake.time),
      format = "%m/%d/%Y %H:%M"
        # MM/DD/YYYY format
    )
  ) %>% # Combine date + time into full timestamps
  group_by(SubjectID, Collection.Date) %>%   
    # Group by subject/day
  fill(reported_wake_time, .direction = "down") %>%  
    # Fill missing wake times so no NAs
  ungroup() %>%
  mutate(
    wake_diff_booklet = as.numeric(booklet_time - reported_wake_time, units = "mins"),
      # Difference between booklet time and wake time
    wake_diff_cap = as.numeric(cap_time - reported_wake_time, units = "mins"),
      # Difference between cap time and wake time
    diff_booklet_cap = as.numeric(booklet_time - cap_time, units = "mins")
      # Difference between booklet time and cap time
  ) # Create new variables as times in mins
```

```{r rq1, include=FALSE}
rq1_agreement <- lmer(wake_diff_booklet ~ wake_diff_cap + (1 | SubjectID), data = analysis_dat)
  # LMM
summary(rq1_agreement)
confint(rq1_agreement, method = "Wald")

rq1_dat <- rq1_agreement@frame %>%
  dplyr::mutate(row_id = as.numeric(rownames(rq1_agreement@frame)))
 # Extract data used in model
rq1_dat$pred_booklet <- predict(rq1_agreement, re.form = NA)
  # Add predictions (population level)

rq1_plot <- ggplot(rq1_dat, aes(x = wake_diff_cap, y = wake_diff_booklet)) +
  geom_point(alpha = 0.6) +
  geom_abline(intercept = 0, slope = 1, color = "red", linetype = "dashed", linewidth = 1) +
  geom_line(aes(y = pred_booklet), color = "blue", linewidth = 1) +
  labs(
    x = "Cap Time Since Wake (min)",
    y = "Booklet Time Since Wake (min)",
    title = "Agreement Between Booklet and Cap"
  ) +
  facet_zoom(xlim = c(0, 50), ylim = c(0, 50), horizontal = FALSE)
    # Zoom in from wake-50 min
```

```{r rq2, include=FALSE}
sleep_adherence <- analysis_dat %>%
  # Assign the correct target for each sample
  mutate(
    target = case_when(
      Collection.Sample == 2 ~ 30,
      Collection.Sample == 4 ~ 600
    )
  ) %>%
  # Reshape so we can evaluate both variables the same way
  pivot_longer(
    cols = c(wake_diff_cap, wake_diff_booklet),
    names_to = "variable_name",
    values_to = "value"
  ) %>%
  # Only keep the samples of interest
  filter(Collection.Sample %in% c(2, 4)) %>%
  group_by(Collection.Sample, variable_name, target) %>%
  summarise(
    n = sum(!is.na(value)),
    n_missing = sum(is.na(value)),
    perc_within_7.5 = mean(abs(value - target) <= 7.5, na.rm = TRUE) * 100,
    perc_within_15  = mean(abs(value - target) <= 15,  na.rm = TRUE) * 100,
    .groups = "drop"
  )

desired_adherence <- analysis_dat %>%
  filter(Collection.Sample %in% c(2, 4)) %>%
  group_by(Collection.Sample) %>%
  summarise(
    n = sum(!is.na(diff_booklet_cap)),
    n_missing = sum(is.na(diff_booklet_cap)),
    perc_within_7.5  = mean(abs(diff_booklet_cap) <= 7.5, na.rm = TRUE) * 100,
    perc_within_15   = mean(abs(diff_booklet_cap) <= 15,  na.rm = TRUE) * 100
  ) # Summarize adherence only for 30min and 10hr samples

# Create a small data frame for the threshold lines
# Threshold line data
thresholds <- data.frame(
  threshold = c(-7.5, 7.5, -15, 15),
  label = c("7.5", "7.5", "15", "15")
)

rq2_plot <- analysis_dat %>%
  pivot_longer(
    cols = c(wake_diff_cap, wake_diff_booklet),
    names_to = "variable_name",
    values_to = "value"
  ) %>%
  # Rename variables for x-axis + legend
  mutate(
    variable_name = recode(variable_name,
                           wake_diff_cap = "Cap",
                           wake_diff_booklet = "Booklet"),
    target = case_when(
      Collection.Sample == 2 ~ 30,
      Collection.Sample == 4 ~ 600
    ),
    diff_from_target = value - target
  ) %>%
  filter(Collection.Sample %in% c(2, 4)) %>%
  ggplot(aes(x = variable_name, y = diff_from_target, fill = variable_name)) +
  
  # Transparent boxplots
  geom_boxplot(alpha = 0.5, outlier.alpha = 0.4) +
  
  # Threshold lines WITH legend
  geom_hline(
    data = thresholds,
    aes(yintercept = threshold, color = label, linetype = label),
    linewidth = 0.8
  ) +
  
  facet_wrap(~ Collection.Sample, scales = "free_y") +
  theme_minimal() +
  labs(
    x = "",
    y = "Difference from Sleep Diary Wake Time (min)",
    title = "Adherence to Measurement Times from Wake Time",
    fill = "Source",
    color = "Threshold",
    linetype = "Threshold"
  ) + 
  coord_cartesian(ylim = c(-50, 150)) +
  scale_color_manual(
    values = c("7.5" = "blue", "15" = "red"),
    labels = c(
      "7.5" = expression("\u00B1" * 7.5 ~ "min"),
      "15"  = expression("\u00B1" * 15 ~ "min")
    )
  ) +
  scale_linetype_manual(
    values = c("7.5" = "dashed", "15" = "dashed"),
    labels = c(
      "7.5" = expression("\u00B1" * 7.5 ~ "min"),
      "15"  = expression("\u00B1" * 15 ~ "min")
    )
  )
rq2_plot
```

```{r rq3 cortisol, include=FALSE}
# Split time into piecewise slopes 
analysis_dat$time_pre30  <- pmin(analysis_dat$wake_diff_booklet, 30)
analysis_dat$time_post30 <- pmax(analysis_dat$wake_diff_booklet - 30, 0)

hist(analysis_dat$Cortisol..nmol.L.); hist(analysis_dat$DHEA..nmol.L.)
analysis_dat$cortisol_log <- log(analysis_dat$Cortisol..nmol.L.)
analysis_dat$DHEA_log <- log(analysis_dat$DHEA..nmol.L.)
hist(analysis_dat$cortisol_log); hist(analysis_dat$DHEA_log)
    # Validating need for log transformation

rq3_cortisol <- lmer(cortisol_log ~ time_pre30 + time_post30 + (1 | SubjectID), data = analysis_dat)
summary(rq3_cortisol)
cort_ci_log <- confint(rq3_cortisol, method = "Wald")
cort_ci_log

# Extract intercept estimate and CI
int_est  <- fixef(rq3_cortisol)["(Intercept)"]
int_lwr  <- cort_ci_log["(Intercept)", 1]
int_upr  <- cort_ci_log["(Intercept)", 2]

# Back-transform
data.frame(
  term = "Wake-time cortisol",
  estimate_nmolL = exp(int_est),
  CI_lower = exp(int_lwr),
  CI_upper = exp(int_upr)
)

pct_change <- function(x) (exp(x) - 1) * 100
  # Helper function for percent change

# Extract slope estimates and CIs
pre_est  <- fixef(rq3_cortisol)["time_pre30"]
pre_lwr  <- cort_ci_log["time_pre30", 1]
pre_upr  <- cort_ci_log["time_pre30", 2]
post_est <- fixef(rq3_cortisol)["time_post30"]
post_lwr <- cort_ci_log["time_post30", 1]
post_upr <- cort_ci_log["time_post30", 2]

data.frame(
  term = c("0–30 min slope", "30 min-10 hr slope"),
  estimate_pct = c(pct_change(pre_est), pct_change(post_est)),
  CI_lower_pct = c(pct_change(pre_lwr), pct_change(post_lwr)),
  CI_upper_pct = c(pct_change(pre_upr), pct_change(post_upr))
) # Back-transform to percent change per minute
  # Multiply by 30 or 60 in reporting

cort_dat <- rq3_cortisol@frame %>%
  mutate(row_id = as.numeric(rownames(rq3_cortisol@frame)))  
  # Extract data used in model
cort_dat$pred_cortisol <- exp(predict(rq3_cortisol, re.form = NA))
  # Add predictions (population level)

analysis_dat <- analysis_dat %>%
  mutate(row_id = row_number())
plot_dat <- cort_dat %>%
  left_join(
    analysis_dat %>% select(row_id, wake_diff_booklet),
    by = "row_id"
  ) # Merge time variable from full dataset

rq3_plot_c <- ggplot(plot_dat, aes(x = wake_diff_booklet, y = exp(cortisol_log))) +
  geom_point(alpha = 0.4) +
  geom_line(aes(y = pred_cortisol), color = "blue", linewidth = 1) +
  geom_vline(xintercept = 30, linetype = "dashed", color = "red") +
  labs(
    x = "Booklet Time Since Wake (min)",
    y = "Cortisol (nmol/L)",
    title = "Predicted Cortisol (Knot at 30 min)"
  ) +
  theme_minimal()
rq3_plot_c
```
```{r rq3 dhea, include=FALSE}
rq3_dhea <- lmer(DHEA_log ~ time_pre30 + time_post30 + (1 | SubjectID), data = analysis_dat)
summary(rq3_dhea)
dhea_ci_log <- confint(rq3_dhea, method = "Wald")
dhea_ci_log

# Extract intercept estimate and CI
int_est  <- fixef(rq3_dhea)["(Intercept)"]
int_lwr  <- dhea_ci_log["(Intercept)", 1]
int_upr  <- dhea_ci_log["(Intercept)", 2]

# Back-transform
data.frame(
  term = "Wake-time DHEA",
  estimate_nmolL = exp(int_est),
  CI_lower = exp(int_lwr),
  CI_upper = exp(int_upr)
)

# Extract slope estimates and CIs
pre_est  <- fixef(rq3_dhea)["time_pre30"]
pre_lwr  <- dhea_ci_log["time_pre30", 1]
pre_upr  <- dhea_ci_log["time_pre30", 2]
post_est <- fixef(rq3_dhea)["time_post30"]
post_lwr <- dhea_ci_log["time_post30", 1]
post_upr <- dhea_ci_log["time_post30", 2]

data.frame(
  term = c("0–30 min slope", "30 min-10 hr slope"),
  estimate_pct = c(pct_change(pre_est), pct_change(post_est)),
  CI_lower_pct = c(pct_change(pre_lwr), pct_change(post_lwr)),
  CI_upper_pct = c(pct_change(pre_upr), pct_change(post_upr))
) # Back-transform to percent change per minute
  # Multiply by 30 or 60 in reporting

dhea_dat <- rq3_dhea@frame %>%
  mutate(row_id = as.numeric(rownames(rq3_dhea@frame)))  
  # Extract data used in model
dhea_dat$pred_dhea <- exp(predict(rq3_dhea, re.form = NA))
  # Add predictions (population level)

analysis_dat <- analysis_dat %>%
  mutate(row_id = row_number())
plot_dat <- dhea_dat %>%
  left_join(
    analysis_dat %>% select(row_id, wake_diff_booklet),
    by = "row_id"
  ) # Merge time variable from full dataset

rq3_plot_d <- ggplot(plot_dat, aes(x = wake_diff_booklet, y = exp(DHEA_log))) +
  geom_point(alpha = 0.4) +
  geom_line(aes(y = pred_dhea), color = "blue", linewidth = 1) +
  geom_vline(xintercept = 30, linetype = "dashed", color = "red") +
  labs(
    x = "Booklet Time Since Wake (min)",
    y = "DHEA (nmol/L)",
    title = "Predicted DHEA (Knot at 30 min)"
  ) +
  theme_minimal()
rq3_plot_d
```
<!-- Title here: -->
# \textit{Timing, Adherence, and Hormone Profiles in a Novel Saliva Device}

<!-- Name here: -->
Ezra Weible[^1]

[^1]: Biostatistics Masters Student, Colorado School of Public Health, CU Anschutz Medical Campus

## Introduction

The investigator provided data from 31 healthy control subjects who collected saliva samples at home using the Saliva Procurement and Integrated Testing (SPIT) booklet, measuring cortisol and DHEA concentrations. Participants recorded sampling times in the booklet, and an electronic monitoring cap independently recorded bottle-opening times. Samples were scheduled at wake, 30 minutes after waking (as reported in sleep diary), before lunch, and 600 minutes (10 hours) after waking. Time of waking and lunch varied by subject.

The scientific questions were: 1) How closely do the booklet reported sampling times agree with electronic cap times?, 2) Do participants adhere to the 30 minute and 10 hour sampling times required by a study protocol?, 3) How cortisol and DHEA change from wake to 30 minutes and from 30 minutes to 10 hours? These translate into the following statistical tasks: 1) Test whether the intercept equals 0 and slope equals 1 in a linear mixed model (LMM) regression of booklet time from wake on cap time from wake., 2) Summarize the proportion samples within $\pm$ 7.5 minutes and $\pm$ 15 minutes at the target times., 3) Test whether cortisol and DHEA increase from wake to 30 minutes and estimate the rate of decline from 30 minutes to 10 hours using a piecewise LMM.

## Methods

All analyses were conducted using R v4.5.1 (2025-06-13 ucrt). Data management included removal of DHEA values at the assay's upper detection limit (5.205 nmol/L) and exclusion of subjects with at least two such values due to possible underlying conditions. Cortisol values exceeding 80 nmol/L were also removed as due to implausibility. Reported sampling times from the booklet and electronic cap and reported wake time from the sleep diary were converted to POSIXct timestamps. For each sample, three variables were computed: 1) booklet time minus wake time, 2) cap time minus wake time, 3) booklet time minus cap time. All differences were stored in minutes. These processed variables were used for all subsequent analyses.

#### Research Question 1:

To evaluate the agreement between participant-reported sampling times in the booklet and reported sampling times from the electronic cap, a LMM was fitted with booklet time from wake as the outcome and cap time from wake as the primary predictor. A random intercept for subject was included to account for repeated measurements within individuals. An LMM was chosen as it accounts for within-subject correlation, unequal observations per participant, and allows for values to be missing at random. The primary goal was to test if booklet and cap times were in perfect agreement, corresponding to two null hypotheses: 1) The intercept is equal to 0, indicating no systematic bias between the two time sources, 2) the slope is equal to 1, indicating a one-to-one relationship between booklet and cap times. A significance threshold of $\alpha$ = 0.05 was used. 

#### Research Question 2:

Adherence to the required 30 minute and 10 hour samples was descriptively assessed. First, the difference between the booklet and cap time for each sample was summarized to quantify reporting consistency. For each sample, the proportion of observations falling within $\pm$ 7.5 minutes and within $\pm$ 15 min at the second and fourth collection sample was calculated, which were intended to correspond to the 30 minute and 10 hour measurements. Second, each recorded time was separately compared to the targeted sampling time relative to the reported wake time in the sleep diary. For each target, the proportion of observations falling within $\pm$ 7.5 minutes and within $\pm$ 15 min was computed. These summaries directly quantify participant compliance and thus the feasibility of the sampling protocol without the need for parametric modeling assumptions. 

#### Research Question 3:

To characterize changes in cortisol and DHEA across the day, piecewise LMM were fitted with a knot at 30 minutes post-wake as reported by the sleep diary. The 30 minute knot was chosen to correspond to the expected patterns of cortisol and DHEA over the course of a day. The primary predictor was booklet time from wake, chosen as it had fewer missing values than cap time and aligned more closely with the sleep diary. Two time variables were constructed, representing minutes from wake to 30 minutes and minutes beyond 30 minutes. The distinct time variables allows for the estimation of separate linear slope for the wake to 30 minute and after 30 minute measurements across the day, to account for the diurnal nature of these hormones. Cortisol and DHEA concentrations were log transformed prior to modeling to improve normality. Each hormone was modeled with the log concentration as the outcome and a random intercept for subject to account for repeated measures.

## Results

*Table 1* summarizes the data across the four daily collection points. Missingness was highest for the cap time measurements, validating the use of booklet times in RQ3. One participant (ID 3037) was excluded due to four DHEA measurements at the assay's upper detection limit.

#### Research Question 1: 

A LMM was used to assess agreement between the recorded booklet and cap sampling times. The slope differed significantly from 0 (p-value < 0.001), indicating that for every one minute increase in cap time, booklet time increased by 0.994 minutes, on average (95\% CI: [0.98, 1.01]). As the 95\% CI overlaps with 1, the cap times reflect a one-to-one relationship with booklet times. However, the intercept differed significantly from 0 (p-value = 0.028), indicating that booklet times were, on average, recorded 6.7 minutes earlier than cap times at wake (95/% CI: [-12.53, -0.88]). Thus, there is systematic bias. This pattern is illustrated in *Figure 1*, where the red dashed line represents perfect agreement and the blue line shows the fitted model. The zoomed-in panel highlights the consistent offset of booklet times relative to cap times. 

#### Research Question 2:

When examining the difference between booklet and cap times for the second and fourth sample of the day, intended to be at 30 minutes and 10 hours post wake, 59-64\% of samples fell within $\pm$ 7.5 minutes and 75-80\% fell within $\pm$ 15 minutes. These times indicate moderate agreement between the two time sources, aligning with the findings from **Research Question 1**.

Adherence to the intended sample times of 30 minutes and 10 hours post wake was evaluated separately for booklet and cap times relative to the reported sleep diary wake time. These findings are visualized in *Figure 2*. Booklet times consistently demonstrated higher adherence than cap times. At the 30 minute sample, 78.6\% of booklet times fell within $\pm$ 7.5 minutes of the target compared to only 52.9\% of cap times. For the $\pm$ 15 minute window, this expanded to 90.5\% of booklet times and 70.6\% of cap times. A similar pattern was observed for the 10 hour sample, though adherence is lower overall, with 48.1\% (booklet) vs. 33.8\% (cap) within $\pm$ 7.5 minutes and 55.8\% (booklet) vs. 41.2\% (cap) within $\pm$ 15 minutes. 

These results, combined with the higher missingness in cap times (*Table 1*), supports the decision to use booklet times as the time variable for modeling cortisol and DHEA in **Research Question 3**. 

#### Research Question 3: 

Piecewise LMMs with a knot at 30 minutes were used to characterize diurnal trajectories of cortisol and DHEA. Both outcomes were modeled on the log scale to improve normality as the original distributions had right skews. 

The estimated intercept for cortisol differed significantly from 0 on the log scale (p-value < 0.001), corresponding to an estimated cortisol concentration at wake of 5.39 nmol/L (95\% CI: [4.41, 6.58]. From wake to 30 minutes, the slope showed a minor increase in cortisol of 0.62% per minute on average (95\% CI: [-0.10, 1.34]), corresponding to an 18.6\% increase over the 30 minutes. However, the slope did not differ significantly from 0 (p-value = 0.094), so we do not have enough evidence to reject reject the null hypothesis that cortisol did not increase from wake to 30 minutes and we cannot conclude that there was a spike during this time. After 30 minutes, cortisol declined significantly (p-value < 0.001). There was an estimated 12\% decrease in cortisol per hour (95\% CI: [11.0, 15.6]). The fitted trajectory is visualized in *Figure 3*, with the knot at 30 minutes indicated by a vertical dashed line.

Unlike cortisol, the estimated intercept for DHEA was not significantly different from 0 on the log scale (p-value = 0.067), corresponding to an estimated DHEA concentration at wake of 1.24 nmol/L (95\% CI: [0.99, 1.56]. From wake to 30 minutes, the slope showed a sharp decrease in DHEA of 1.76% per minute on average (95\% CI: [-2.35, -1.17]), corresponding to an 52.8\% decrease over the 30 minutes (p-value < 0.001). After 30 minutes, DHEA continued to decline at a slower but still significant rate (p-value < 0.001). There was an estimated 9.2\% decrease in DHEA per hour (95\% CI: [-7.44, -11.04]). The fitted trajectory is visualized in *Figure 4*.

Overall, both hormones showed an expected decline after 30 minutes. However, neither had the anticipated spike from wake to 30 minutes, with cortisol having a mild increase and DHEA having a sharp drop. Although these results support the use of a piecewise model with a knot at 30 minutes, they do not align with the anticipated clinical results for these hormones, which could indicate issues with the novel method. 

## Discussion

This study evaluated the feasibility and accuracy of a novel at‑home saliva collection method using the SPIT booklet, with a focus on agreement between reported and electronically recorded sampling times, adherence to protocol‑specified sampling windows, and characterization of diurnal cortisol and DHEA patterns. Together, the findings provide insight into both participant behavior and the performance of the collection device in a naturalistic setting.

Several limitations should be considered. The study lacked a true ground‑truth measure of sampling time, making it impossible to determine whether discrepancies reflected reporting error, device malfunction, or participant behavior. The sample size was modest, which may limit generalizability and reduce power to detect subtle effects. Assay constraints required removal of DHEA values at the upper detection limit, and one participant was excluded for repeated censoring. The electronic cap did not function consistently, contributing to missingness and limiting its utility as a reference standard. Finally, at‑home collection introduces natural variability in participant routines and environmental conditions. Despite these limitations, the findings support the feasibility of the SPIT booklet and provide guidance for improving timing accuracy in future at‑home endocrine studies.

## Appendix

Directory: [Github](https://github.com/eweible/BIOS6624.git)

File name: P0/Code/P0_Report.Rmd

#### Tables and Figures:

```{r echo = FALSE, tab.cap="Table 1"}
tab1 <- analysis_dat %>%
  select(
    Collection.Sample,
    wake_diff_cap,
    wake_diff_booklet,
    diff_booklet_cap,
    Cortisol..nmol.L.,
    DHEA..nmol.L.
  ) %>%
  tbl_summary(
    by = Collection.Sample,
    missing = "ifany",
    missing_text = "Missing",
      # Rename missing
    statistic = list(
      all_continuous() ~ "{mean} ({sd})",
      c(Cortisol..nmol.L., DHEA..nmol.L.) ~ "{median} [{p25}, {p75}]"
        # Median/IQR for cortisol and DHEA because of skew
    ),
    digits = everything() ~ 2,
    label = list(
      wake_diff_cap ~ "Cap – Wake Time (min)",
      wake_diff_booklet ~ "Booklet – Wake Time (min)",
      diff_booklet_cap ~ "Booklet – Cap Time (min)",
      Cortisol..nmol.L. ~ "Cortisol (nmol/L)",
      DHEA..nmol.L. ~ "DHEA (nmol/L)"
    ) # Cleaner labels
  ) %>%
  modify_header(label = "**Variable**") %>%
  modify_caption("**Summary of Missingness and Key Variables by Collection Sample**")

tab1
```

```{r, echo=FALSE, fig.cap="Research Question 1 Agreement between reported booklet time since wake and reported cap time since wake."}
rq1_plot
```

```{r, echo=FALSE, fig.cap="Research Question 2 Adherence to repoting windows.", warning=FALSE}
rq2_plot
```

```{r, echo=FALSE, fig.cap="Research Question 3 Predicted trend of cortisol over time since wake, reported from booklet."}
rq3_plot_c
```

```{r, echo=FALSE, fig.cap="Research Question 3 Predicted trend of DHEA over time since wake, reported from booklet."}
rq3_plot_d
```

