---
output:
  bookdown::pdf_document2:
    toc: false
    number_sections: false
documentclass: article
geometry: margin=1in
fontsize: 12pt
linestretch: 2
fig.cap: true
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
library(dplyr)
library(tidyr)
library(lmerTest)
library(ggplot2)
library(ggforce)
raw_dat <- read.csv(here::here("P0/DataRaw", "Project0_Clean_v2.csv"))
  # Read in dataset 
```

```{r data_manip, include=FALSE}
analysis_dat <- raw_dat %>%
  select(
    -Booklet..Sample.interval,
    -Booklet..Sample.interval.Decimal.Time..mins.,
    -MEMs..Sample.interval,
    -MEMs..Sample.interval.Decimal.Time..mins.,
    -Cortisol..ug.dl.,
    -DHEA..pg.dl.
  ) %>% # Remove unwanted columns
  
  group_by(SubjectID) %>%
  mutate(
    dhea_upper_hits = sum(DHEA..nmol.L. == 5.205, na.rm = TRUE)
  ) %>%
  ungroup() %>% # Identify number of DHEA values at the detection limit (5.205)
  { 
    removed_info <<- distinct(.[.$dhea_upper_hits > 2, c("SubjectID", "dhea_upper_hits")])
    . # Return to data
  } %>% # Save IDs & number of hits of participants to be removed
  filter(!SubjectID %in% removed_info) %>%
    # Remove participants with >2 detection limit hits across all their rows
  
  mutate(
    Cortisol..nmol.L. = ifelse(Cortisol..nmol.L. > 80, NA, Cortisol..nmol.L.),
      # Remove cortisol > 80 nmol/L
    DHEA..nmol.L. = ifelse(DHEA..nmol.L. == 5.205, NA, DHEA..nmol.L.)
      # Remove DHEA at the detection limit
  ) %>% 
  select(-dhea_upper_hits)
    # Drop helper column

# # Print about removal to share with investigator
# cat("Number of participants removed:", nrow(removed_info), "\n")
# cat("Removed participant ID(s):", removed_info$SubjectID, "\n")
# cat("Number of DHEA values at detection limit:", removed_info$dhea_upper_hits)

analysis_dat <- analysis_dat %>%
  mutate(
    Booket..Clock.Time = na_if(trimws(Booket..Clock.Time), ""),
    MEMs..Clock.Time = na_if(trimws(MEMs..Clock.Time), ""),
    Sleep.Diary.reported.wake.time = na_if(trimws(Sleep.Diary.reported.wake.time), "")
  ) %>% # Convert empty strings to NAs and trim leading spaces
  mutate(
    booklet_time = as.POSIXct(
      paste(Collection.Date, Booket..Clock.Time),
      format = "%m/%d/%Y %H:%M"
        # MM/DD/YYYY format
    ),
    cap_time = as.POSIXct(
      paste(Collection.Date, MEMs..Clock.Time),
      format = "%m/%d/%Y %H:%M"
        # MM/DD/YYYY format
    ),
    reported_wake_time = as.POSIXct(
      paste(Collection.Date, Sleep.Diary.reported.wake.time),
      format = "%m/%d/%Y %H:%M"
        # MM/DD/YYYY format
    )
  ) %>% # Combine date + time into full timestamps
  group_by(SubjectID, Collection.Date) %>%   
    # Group by subject/day
  fill(reported_wake_time, .direction = "down") %>%  
    # Fill missing wake times so no NAs
  ungroup() %>%
  mutate(
    wake_diff_booklet = as.numeric(booklet_time - reported_wake_time, units = "mins"),
      # Difference between booklet time and wake time
    wake_diff_cap = as.numeric(cap_time - reported_wake_time, units = "mins"),
      # Difference between cap time and wake time
    diff_booklet_cap = as.numeric(booklet_time - cap_time, units = "mins")
      # Difference between booklet time and cap time
  ) # Create new variables as times in mins
```

```{r rq1, include=FALSE}
rq1_agreement <- lmer(wake_diff_booklet ~ wake_diff_cap + (1 | SubjectID), data = analysis_dat)
  # LMM
summary(rq1_agreement)

car::linearHypothesis(rq1_agreement, "wake_diff_cap = 1")
  # Test if slope is equal to 1

rq1_dat <- rq1_agreement@frame %>%
  dplyr::mutate(row_id = as.numeric(rownames(rq1_agreement@frame)))
 # Extract data used in model
rq1_dat$pred_booklet <- predict(rq1_agreement, re.form = NA)
  # Add predictions (population level)

rq1_plot <- ggplot(rq1_dat, aes(x = wake_diff_cap, y = wake_diff_booklet)) +
  geom_point(alpha = 0.6) +
  geom_abline(intercept = 0, slope = 1, color = "red", linetype = "dashed", linewidth = 1) +
  geom_line(aes(y = pred_booklet), color = "blue", linewidth = 1) +
  labs(
    x = "Cap Time Since Wake (min)",
    y = "Booklet Time Since Wake (min)",
    title = "Agreement Between Booklet and Cap"
  ) +
  facet_zoom(xlim = c(0, 50), ylim = c(0, 50), horizontal = FALSE)
    # Zoom in from wake-50 min
```

```{r rq2, include=FALSE}
sleep_adherence <- analysis_dat %>%
  # Assign the correct target for each sample
  mutate(
    target = case_when(
      Collection.Sample == 2 ~ 30,
      Collection.Sample == 4 ~ 600
    )
  ) %>%
  # Reshape so we can evaluate both variables the same way
  pivot_longer(
    cols = c(wake_diff_cap, wake_diff_booklet),
    names_to = "variable_name",
    values_to = "value"
  ) %>%
  # Only keep the samples of interest
  filter(Collection.Sample %in% c(2, 4)) %>%
  group_by(Collection.Sample, variable_name, target) %>%
  summarise(
    n = sum(!is.na(value)),
    n_missing = sum(is.na(value)),
    perc_within_7.5 = mean(abs(value - target) <= 7.5, na.rm = TRUE) * 100,
    perc_within_15  = mean(abs(value - target) <= 15,  na.rm = TRUE) * 100,
    .groups = "drop"
  )

desired_adherence <- analysis_dat %>%
  filter(Collection.Sample %in% c(2, 4)) %>%
  group_by(Collection.Sample) %>%
  summarise(
    n = sum(!is.na(diff_booklet_cap)),
    n_missing = sum(is.na(diff_booklet_cap)),
    perc_within_7.5  = mean(abs(diff_booklet_cap) <= 7.5, na.rm = TRUE) * 100,
    perc_within_15   = mean(abs(diff_booklet_cap) <= 15,  na.rm = TRUE) * 100
  ) # Summarize adherence only for 30min and 10hr samples

# Create a small data frame for the threshold lines
# Threshold line data
thresholds <- data.frame(
  threshold = c(-7.5, 7.5, -15, 15),
  label = c("7.5", "7.5", "15", "15")
)

rq2_plot <- analysis_dat %>%
  pivot_longer(
    cols = c(wake_diff_cap, wake_diff_booklet),
    names_to = "variable_name",
    values_to = "value"
  ) %>%
  # Rename variables for x-axis + legend
  mutate(
    variable_name = recode(variable_name,
                           wake_diff_cap = "Cap",
                           wake_diff_booklet = "Booklet"),
    target = case_when(
      Collection.Sample == 2 ~ 30,
      Collection.Sample == 4 ~ 600
    ),
    diff_from_target = value - target
  ) %>%
  filter(Collection.Sample %in% c(2, 4)) %>%
  ggplot(aes(x = variable_name, y = diff_from_target, fill = variable_name)) +
  
  # Transparent boxplots
  geom_boxplot(alpha = 0.5, outlier.alpha = 0.4) +
  
  # Threshold lines WITH legend
  geom_hline(
    data = thresholds,
    aes(yintercept = threshold, color = label, linetype = label),
    linewidth = 0.8
  ) +
  
  facet_wrap(~ Collection.Sample, scales = "free_y") +
  theme_minimal() +
  labs(
    x = "",
    y = "Difference from Sleep Diary Wake Time (min)",
    title = "Adherence to Measurement Times from Wake Time",
    fill = "Source",
    color = "Threshold",
    linetype = "Threshold"
  ) + 
  coord_cartesian(ylim = c(-50, 150)) +
  scale_color_manual(
    values = c("7.5" = "blue", "15" = "red"),
    labels = c(
      "7.5" = expression("\u00B1" * 7.5 ~ "min"),
      "15"  = expression("\u00B1" * 15 ~ "min")
    )
  ) +
  scale_linetype_manual(
    values = c("7.5" = "dashed", "15" = "dashed"),
    labels = c(
      "7.5" = expression("\u00B1" * 7.5 ~ "min"),
      "15"  = expression("\u00B1" * 15 ~ "min")
    )
  )
rq2_plot
```

```{r rq3 cortisol, include=FALSE}
# Split time into piecewise slopes 
analysis_dat$time_pre30  <- pmin(analysis_dat$wake_diff_booklet, 30)
analysis_dat$time_post30 <- pmax(analysis_dat$wake_diff_booklet - 30, 0)

hist(analysis_dat$Cortisol..nmol.L.); hist(analysis_dat$DHEA..nmol.L.)
analysis_dat$cortisol_log <- log(analysis_dat$Cortisol..nmol.L.)
analysis_dat$DHEA_log <- log(analysis_dat$DHEA..nmol.L.)
hist(analysis_dat$cortisol_log); hist(analysis_dat$DHEA_log)
    # Validating need for log transformation

rq3_cortisol <- lmer(cortisol_log ~ time_pre30 + time_post30 + (1 | SubjectID), data = analysis_dat)
summary(rq3_cortisol)

cort_dat <- rq3_cortisol@frame %>%
  mutate(row_id = as.numeric(rownames(rq3_cortisol@frame)))  
  # Extract data used in model
cort_dat$pred_cortisol <- exp(predict(rq3_cortisol, re.form = NA))
  # Add predictions (population level)

analysis_dat <- analysis_dat %>%
  mutate(row_id = row_number())
plot_dat <- cort_dat %>%
  left_join(
    analysis_dat %>% select(row_id, wake_diff_booklet),
    by = "row_id"
  ) # Merge time variable from full dataset

rq3_plot_c <- ggplot(plot_dat, aes(x = wake_diff_booklet, y = exp(cortisol_log))) +
  geom_point(alpha = 0.4) +
  geom_line(aes(y = pred_cortisol), color = "blue", linewidth = 1) +
  geom_vline(xintercept = 30, linetype = "dashed", color = "red") +
  labs(
    x = "Booklet Time Since Wake (min)",
    y = "Cortisol (nmol/L)",
    title = "Predicted Cortisol (Knot at 30 min)"
  ) +
  theme_minimal()
rq3_plot_c
```
```{r rq3 dhea, include=FALSE}
rq3_dhea <- lmer(DHEA_log ~ time_pre30 + time_post30 + (1 | SubjectID), data = analysis_dat)
summary(rq3_dhea)

dhea_dat <- rq3_dhea@frame %>%
  mutate(row_id = as.numeric(rownames(rq3_dhea@frame)))  
  # Extract data used in model
dhea_dat$pred_dhea <- exp(predict(rq3_dhea, re.form = NA))
  # Add predictions (population level)

analysis_dat <- analysis_dat %>%
  mutate(row_id = row_number())
plot_dat <- dhea_dat %>%
  left_join(
    analysis_dat %>% select(row_id, wake_diff_booklet),
    by = "row_id"
  ) # Merge time variable from full dataset

rq3_plot_d <- ggplot(plot_dat, aes(x = wake_diff_booklet, y = exp(DHEA_log))) +
  geom_point(alpha = 0.4) +
  geom_line(aes(y = pred_dhea), color = "blue", linewidth = 1) +
  geom_vline(xintercept = 30, linetype = "dashed", color = "red") +
  labs(
    x = "Booklet Time Since Wake (min)",
    y = "DHEA (nmol/L)",
    title = "Predicted DHEA (Knot at 30 min)"
  ) +
  theme_minimal()
rq3_plot_d
```
<!-- Title here: -->
# \textit{Timing, Adherence, and Hormone Profiles in a Novel Saliva Device}

<!-- Name here: -->
Ezra Weible[^1]

[^1]: Biostatistics Masters Student, Colorado School of Public Health, CU Anschutz Medical Campus

## Introduction

Cortisol and DHEA are key hormones that show diurnal patterns over the course of a day, and changes to these patterns can be used to identify health conditions. However, collection of these trajectories requires accurate timing of saliva samples, particularly when measuring participants in typical setting such as at home compared to laboratory environments. This project aims to investigate the feasibility of a novel method to collect saliva samples at at-home. 31 healthy control subjects collected saliva samples using the Saliva Procurement and Integrated Testing (SPIT) booklet, which provided collection strips on filter paper. Participants recorded the time of each sample in the booklet, and the electronic monitoring cap on the storage bottle independently recorded the time it was opened. The participants took samples four times a day for three days scheduled at time of wake (identified by subject's reported wake time in their sleep diary), 30 minutes after waking, before lunch, and 600 minutes (10 hours) after waking. Time of waking and lunch were allowed to vary by subject. 

This study aimed to answer three scientific questions: 1) How closely do the participant-reported saliva collection times in the booklet correspond to the electronically recorded times from the electronic monitoring cap?, 2) To what extent do participants adhere to the 30 minute and 10 hour sampling times required by a study protocol?, 3) How do salivary cortisol and DHEA concentrations change over the day, specifically from waking to 30 minutes and from 30 minutes to 10 hours? 

The purpose of this report is to develop an analysis plan and report on the results of these scientific questions, which translate into the following statistical hypotheses or tasks: 1) Null Hypothesis: Intercept = 0 and slope = 1 in a linear mixed model (LMM) regression of book time from wake on cap time from wake with subject-level random intercepts. Alternative: The intercept $\neq$ 0 or the slope $\neq$ 1., 2) Descriptively, what proportion of subjects fall within $\pm$ 7.5 minutes (good adherence) and $\pm$ 15 minutes (adequate adherence) at the 30 minute and 10 hour measurements?, 3) Null Hypothesis: The change in cortisol from wake to 30 minutes = 0 and the change in DHEA from wake to 30 minutes = 0. Alternative: The change in cortisol from wake to 30 minutes > 0 and the change in DHEA from wake to 30 minutes > 0. 

## Methods

All analyses were conducted using R v4.5.1 (2025-06-13 ucrt). Prior to statistical modeling, data management was performed to ensure data quality and remove clinically implausible. As the DHEA assay had an upper detection limit of 5.205 nmol/L, all DHEA measurements at this limitewere excluded. In addition, subjects who had at least two DHEA measurements at this upper detection limit were excluded, as it may indicate underlying endocrine conditions or assay errors. Cortisol values exceeding 80 nmol/L were also removed as they are considered clinically implausible and are likely measurement error. 

Reported sampling times from the booklet and electronic cap and reported wake time from the sleep diary were converted to POSIXct timestamps. For each sample, three variables were computed: 1) booklet time minus wake time, 2) cap time minus wake time, 3) booklet time minus cap time. All differences were stored in minutes. These processed variables were used for all subsequent analyses.

#### Research Question 1:

To evaluate the agreement between participant-reported sampling times in the booklet and reported sampling times from the electronic cap, a LMM was fitted with booklet time from wake as the outcome and cap time from wake as the primary predictor. A random intercept for subject was included to account for repeated measurements within individuals. An LMM was chosen as it accounts for within-subject correlation, unequal observations per participant, and allows for values to be missing at random. The primary goal was to test if booklet and cap times were in perfect agreement, corresponding to two null hypotheses: 1) The intercept is equal to 0, indicating no systematic bias between the two time sources, 2) the slope is equal to 1, indicating a one-to-one relationship between booklet and cap times. These hypotheses were tested using a significance threshold of $\alpha$ = 0.05. 

#### Research Question 2:

Adherence to the required 30 minute and 10 hour samples was descriptively assessed. First, the difference between the booklet and cap time for each sample was summarized to quantify reporting consistency. For each sample, the proportion of observations falling within $\pm$ 7.5 minutes (good adherence) and within $\pm$ 15 min (adequate adherence) at the second and fourth collection sample was calculated, which were intended to correspond to the 30 minute and 10 hour measurements. Second, each recorded time was separately compared to the targeted sampling time relative to the reported wake time in the sleep diary. For each target, the proportion of observations falling within $\pm$ 7.5 minutes and within $\pm$ 15 min was similarly computed. Descriptive summaries were selected as they directly quantify participant compliance and thus the feasibility of the sampling protocol without the need for parametric assumptions. 

#### Research Question 3:

To characterize changes in cortisol and DHEA across the day, piecewise LMM were fitted with a knot at 30 minutes post-wake as reported by the sleep diary. The 30 minute knot was chosen to correspond to the expected patterns of cortisol and DHEA over the course of a day. The primary predictor was booklet time from wake, chosen as it had fewer missing values than cap time and aligned more closely with the sleep diary. Two time variables were constructed, representing minutes from wake to 30 minutes and minutes beyond 30 minutes. The distinct time variables allows for the estimation of separate linear slope for the wake to 30 minute and after 30 minute measurements across the day, to account for the diurnal nature of these hormones. Cortisol and DHEA concentrations were log transformed prior to modeling to improve normality. Each hormone was modeled with the log concentration as the outcome and a random intercept for subject to account for repeated measures.

## Results

# TODO - update table 1 and summary 

*Table 1* summarizes the completeness of the sampling data across the four daily collection points. The majority of the missing data was for the cap time measurements, so when choosing between cap and booklet, booklet was selected. 

Only one participant (ID 3037) was excluded due to having four DHEA measurements at the assay's upper detection limit (5.205 nmol/L), suggesting potential underlying abnormalities or assay issues.

#### Research Question 1: 

A LMM was used to assess agreement between the recorded booklet and electronic cap sampling times. The fixed effect estimates indicated that both the intercept and slope differed significantly from 0. INTERCEPT PT EST+CI+PVAL. SLOPE PT EST+CI+PVAL. The variance for subject-specific intercepts in the random effects was modest (SD = 6.11), but the residual variance was large (SD = 31.97), indicating high within-person variability in the timing differences. 

To evaluate whether the slope differed from the hypothesized value of 1, an additional linear hypothesis test was conducted. There was no evidence that the slope deviated from 1 (p-value = 0.389). This suggests that changes in cap time were reflected nearly one-to-one in booklet time. However, the intercept differed significantly from 0, indicating systematic bias. On average, booklet times were 6.7 minutes earlier than cap times at wake time. This pattern is illustrated in *Figure 1*, where the red dashed line represents perfect agreement and the blue line shows the fitted model. The zoomed-in panel highlights the consistent offset of booklet times relative to cap times. 

#### Research Question 2:

When examining the difference between booklet and cap times for the second and fourth sample of the day, intended to be at 30 minutes and 10 hours post wake, 59-64\% of samples fell within $\pm$ 7.5 minutes and 75-80\% fell within $\pm$ 15 minutes. These times indicate moderate agreement between the two time sources, aligning with the findings from **Research Question 1**.

Adherence to the intended sample times of 30 minutes and 10 hours post wake was evaluated separately for booklet and cap times relative to the reported sleep diary wake time. These findings are visualized in *Figure 2*. Booklet times consistently demonstrated higher adherence than cap times. At the 30 minute sample, 78.6\% of booklet times fell within $\pm$ 7.5 minutes of the target compared to only 52.9\% of cap times. For the $\pm$ 15 minute window, this expanded to 90.5\% of booklet times and 70.6\% of cap times. A similar pattern was observed for the 10 hour sample, though adherence is lower overall, with 48.1\% (booklet) vs. 33.8\% (cap) within $\pm$ 7.5 minutes and 55.8\% (booklet) vs. 41.2\% (cap) within $\pm$ 15 minutes. 

These results, combined with the higher missingness in cap times (*Table 1*), supports the decision to use booklet times as the time variable for modeling cortisol and DHEA in **Research Question 3**. 


#### Research Question 3: 

Piecewise LMMs with a knot at 30 minutes were used to characterize diurnal trajectories of cortisol and DHEA. Both outcomes were modeled on the log scale to improve normality as the original distributions had right skews. 

The estimated intercept for cortisol_log was 1.684 (SE = 0.102, p < 0.001), corresponding to an expected cortisol concentration of approximately 5.39 nmol/L at wake time. The slope from 0–30 minutes was positive (0.0061 per minute, SE = 0.0036), but only marginally significant (p = 0.094), indicating a modest and statistically uncertain increase in cortisol during the expected awakening response period. After 30 minutes, cortisol declined significantly at a rate of –0.00223 per minute (SE = 0.00019, p < 0.001). This corresponds to an approximate 12% decrease per hour after the 30‑minute mark. The fitted trajectory is shown in *Figure 3*, with the knot at 30 minutes indicated by a vertical dashed line.

For DHEA, the intercept was 0.219 (SE = 0.117, p = 0.067), corresponding to an estimated wake‑time concentration of 1.24 nmol/L. Unlike cortisol, DHEA showed a significant decrease during the first 30 minutes (slope = –0.0178 per minute, SE = 0.0031, p < 0.001). After 30 minutes, DHEA continued to decline at a slower but still significant rate (–0.00154 per minute, SE = 0.00015, p < 0.001). The modeled trajectory is shown in *Figure 4*.

# TODO - update these two paragraphs to be better and include CI

Overall, both hormones showed an expected decline after 30 minutes. However, neither had the anticipated spike from wake to 30 minutes, with cortisol having a mild increase and DHEA having a sharp drop. Although these results support the use of a piecewise model with a knot at 30 minutes, they do not align with the anticipated clinical results for these hormones, which could indicate issues with the novel method. 

## Discussion

This study evaluated the feasibility and accuracy of a novel at‑home saliva collection method using the SPIT booklet, with a focus on agreement between reported and electronically recorded sampling times, adherence to protocol‑specified sampling windows, and characterization of diurnal cortisol and DHEA patterns. Together, the findings provide insight into both participant behavior and the performance of the collection device in a naturalistic setting.

Several limitations should be considered. The study lacked a true ground‑truth measure of sampling time, making it impossible to determine whether discrepancies reflected reporting error, device malfunction, or participant behavior. The sample size was modest, which may limit generalizability and reduce power to detect subtle effects. Assay constraints required removal of DHEA values at the upper detection limit, and one participant was excluded for repeated censoring. The electronic cap did not function consistently, contributing to missingness and limiting its utility as a reference standard. Finally, at‑home collection introduces natural variability in participant routines and environmental conditions. Despite these limitations, the findings support the feasibility of the SPIT booklet and provide guidance for improving timing accuracy in future at‑home endocrine studies.

## Appendix

Directory: [Github](https://github.com/eweible/BIOS6624.git)

File name: P0/Code/P0_Report.Rmd

#### Tables and Figures:

```{r echo = FALSE, tab.cap="Table 1"}
library(arsenal)

tab <- tableby(
  Collection.Sample ~ wake_diff_cap + wake_diff_booklet + diff_booklet_cap +
    Cortisol..nmol.L. + DHEA..nmol.L.,
  data = analysis_dat
)

summary(tab)
```

```{r, echo=FALSE, fig.cap="Research Question 1 Agreement between reported booklet time since wake and reported cap time since wake."}
rq1_plot
```

```{r, echo=FALSE, fig.cap="Research Question 2 Adherence to repoting windows.", warning=FALSE}
rq2_plot
```

```{r, echo=FALSE, fig.cap="Research Question 3 Predicted trend of cortisol over time since wake, reported from booklet."}
rq3_plot_c
```

```{r, echo=FALSE, fig.cap="Research Question 3 Predicted trend of DHEA over time since wake, reported from booklet."}
rq3_plot_d
```

