---
output:
  bookdown::pdf_document2:
    toc: false
    number_sections: false
documentclass: article
geometry: margin=1in
fontsize: 12pt
linestretch: 2
fig.cap: true
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
library(dplyr)
library(tidyr)
library(lmerTest)
library(ggplot2)
library(ggforce)
raw_dat <- read.csv(here::here("P0/DataRaw", "Project0_Clean_v2.csv"))
  # Read in dataset 
```

```{r data_manip, include=FALSE}
analysis_dat <- raw_dat %>%
  select(
    -Booklet..Sample.interval,
    -Booklet..Sample.interval.Decimal.Time..mins.,
    -MEMs..Sample.interval,
    -MEMs..Sample.interval.Decimal.Time..mins.,
    -Cortisol..ug.dl.,
    -DHEA..pg.dl.
  ) %>% # Remove unwanted columns
  
  group_by(SubjectID) %>%
  mutate(
    dhea_upper_hits = sum(DHEA..nmol.L. == 5.205, na.rm = TRUE)
  ) %>%
  ungroup() %>% # Identify number of DHEA values at the detection limit (5.205)
  { 
    removed_info <<- distinct(.[.$dhea_upper_hits > 2, c("SubjectID", "dhea_upper_hits")])
    . # Return to data
  } %>% # Save IDs & number of hits of participants to be removed
  filter(!SubjectID %in% removed_info) %>%
    # Remove participants with >2 detection limit hits across all their rows
  
  mutate(
    Cortisol..nmol.L. = ifelse(Cortisol..nmol.L. > 80, NA, Cortisol..nmol.L.),
      # Remove cortisol > 80 nmol/L
    DHEA..nmol.L. = ifelse(DHEA..nmol.L. == 5.205, NA, DHEA..nmol.L.)
      # Remove DHEA at the detection limit
  ) %>% 
  select(-dhea_upper_hits)
    # Drop helper column
analysis_dat <- analysis_dat %>%
  mutate(
    Booket..Clock.Time = na_if(trimws(Booket..Clock.Time), ""),
    MEMs..Clock.Time = na_if(trimws(MEMs..Clock.Time), ""),
    Sleep.Diary.reported.wake.time = na_if(trimws(Sleep.Diary.reported.wake.time), "")
  ) %>% # Convert empty strings to NAs and trim leading spaces
  mutate(
    booklet_time = as.POSIXct(
      paste(Collection.Date, Booket..Clock.Time),
      format = "%m/%d/%Y %H:%M"
        # MM/DD/YYYY format
    ),
    cap_time = as.POSIXct(
      paste(Collection.Date, MEMs..Clock.Time),
      format = "%m/%d/%Y %H:%M"
        # MM/DD/YYYY format
    ),
    reported_wake_time = as.POSIXct(
      paste(Collection.Date, Sleep.Diary.reported.wake.time),
      format = "%m/%d/%Y %H:%M"
        # MM/DD/YYYY format
    )
  ) %>% # Combine date + time into full timestamps
  group_by(SubjectID, Collection.Date) %>%   
    # Group by subject/day
  fill(reported_wake_time, .direction = "down") %>%  
    # Fill missing wake times so no NAs
  ungroup() %>%
  mutate(
    wake_diff_booklet = as.numeric(booklet_time - reported_wake_time, units = "mins"),
      # Difference between booklet time and wake time
    wake_diff_cap = as.numeric(cap_time - reported_wake_time, units = "mins"),
      # Difference between cap time and wake time
    diff_booklet_cap = as.numeric(booklet_time - cap_time, units = "mins")
      # Difference between booklet time and cap time
  ) # Create new variables as times in mins
```

```{r rq1, include=FALSE}
rq1_agreement <- lmer(wake_diff_booklet ~ wake_diff_cap + (1 | SubjectID), data = analysis_dat)
  # LMM
summary(rq1_agreement)

car::linearHypothesis(rq1_agreement, "wake_diff_cap = 1")
  # Test if slope is equal to 1

rq1_dat <- rq1_agreement@frame %>%
  dplyr::mutate(row_id = as.numeric(rownames(rq1_agreement@frame)))
 # Extract data used in model
rq1_dat$pred_booklet <- predict(rq1_agreement, re.form = NA)
  # Add predictions (population level)

rq1_plot <- ggplot(rq1_dat, aes(x = wake_diff_cap, y = wake_diff_booklet)) +
  geom_point(alpha = 0.6) +
  geom_abline(intercept = 0, slope = 1, color = "red", linetype = "dashed", linewidth = 1) +
  geom_line(aes(y = pred_booklet), color = "blue", linewidth = 1) +
  labs(
    x = "Cap Time Since Wake (min)",
    y = "Booklet Time Since Wake (min)",
    title = "Agreement Between Booklet and Cap"
  ) +
  facet_zoom(xlim = c(0, 50), ylim = c(0, 50), horizontal = FALSE)
    # Zoom in from wake-50 min
```

```{r rq2, include=FALSE}
analysis_dat %>%
  summarise(
    n = sum(!is.na(diff_booklet_cap)),
    n_missing = sum(is.na(diff_booklet_cap)),
    perc_within_7.5  = mean(abs(diff_booklet_cap) <= 7.5, na.rm = TRUE) * 100,
    perc_within_15   = mean(abs(diff_booklet_cap) <= 15,  na.rm = TRUE) * 100
  ) #Summarize adherence overall 

analysis_dat %>%
  filter(Collection.Sample %in% c(2, 4)) %>%
  group_by(Collection.Sample) %>%
  summarise(
    n = sum(!is.na(diff_booklet_cap)),
    n_missing = sum(is.na(diff_booklet_cap)),
    perc_within_7.5  = mean(abs(diff_booklet_cap) <= 7.5, na.rm = TRUE) * 100,
    perc_within_15   = mean(abs(diff_booklet_cap) <= 15,  na.rm = TRUE) * 100
  ) # Summarize adherence only for 30min and 10hr samples
```

# TODO - update rq2 to be a nice table and insert table below

```{r rq3 cortisol, include=FALSE}
# Split time into piecewise slopes 
analysis_dat$time_pre30  <- pmin(analysis_dat$wake_diff_booklet, 30)
analysis_dat$time_post30 <- pmax(analysis_dat$wake_diff_booklet - 30, 0)

rq3_cortisol <- lmer(Cortisol..nmol.L. ~ time_pre30 + time_post30 + (1 | SubjectID), data = analysis_dat)
summary(rq3_cortisol)

cort_dat <- rq3_cortisol@frame %>%
  mutate(row_id = as.numeric(rownames(rq3_cortisol@frame)))  
  # Extract data used in model
cort_dat$pred_cortisol <- predict(rq3_cortisol, re.form = NA)
  # Add predictions (population level)

analysis_dat <- analysis_dat %>%
  mutate(row_id = row_number())
plot_dat <- cort_dat %>%
  left_join(
    analysis_dat %>% select(row_id, wake_diff_booklet),
    by = "row_id"
  ) # Merge time variable from full dataset

rq3_plot_c <- ggplot(plot_dat, aes(x = wake_diff_booklet, y = Cortisol..nmol.L.)) +
  geom_point(alpha = 0.4) +
  geom_line(aes(y = pred_cortisol), color = "blue", linewidth = 1) +
  geom_vline(xintercept = 30, linetype = "dashed", color = "red") +
  labs(
    x = "Booklet Time Since Wake (min)",
    y = "Cortisol (nmol/L)",
    title = "Predicted Cortisol (Knot at 30 min)"
  ) +
  theme_minimal()
```
```{r rq3 dhea, include=FALSE}
rq3_dhea <- lmer(DHEA..nmol.L. ~ time_pre30 + time_post30 + (1 | SubjectID), data = analysis_dat)
summary(rq3_dhea)

dhea_dat <- rq3_dhea@frame %>%
  mutate(row_id = as.numeric(rownames(rq3_dhea@frame)))  
  # Extract data used in model
dhea_dat$pred_dhea <- predict(rq3_dhea, re.form = NA)
  # Add predictions (population level)

analysis_dat <- analysis_dat %>%
  mutate(row_id = row_number())
plot_dat <- dhea_dat %>%
  left_join(
    analysis_dat %>% select(row_id, wake_diff_booklet),
    by = "row_id"
  ) # Merge time variable from full dataset

rq3_plot_d <- ggplot(plot_dat, aes(x = wake_diff_booklet, y = DHEA..nmol.L.)) +
  geom_point(alpha = 0.4) +
  geom_line(aes(y = pred_dhea), color = "blue", linewidth = 1) +
  geom_vline(xintercept = 30, linetype = "dashed", color = "red") +
  labs(
    x = "Booklet Time Since Wake (min)",
    y = "DHEA (nmol/L)",
    title = "Predicted DHEA (Knot at 30 min)"
  ) +
  theme_minimal()
```
<!-- Title here: -->
# \textit{An Instigation into Time Agreement, Subject Adherence, and Patterns of Cortisol and DHEA in Novel Saliva Collection Device}

<!-- Name here: -->
Ezra Weible[^1]

[^1]: Biostatistics Masters Student, Colorado School of Public Health, University of Colorado Anschutz Medical Campus, Aurora, CO

## Introduction

0.5 pages; describe data and scientific hypotheses of interest. Rephrase scientific hypotheses into statistical hypotheses. 

## Methods

2 pages; past tense; no results; brief description of data management and cleaning. In depth description of analysis methods. Justify and explain data analysis approach. 

## Results

1-1.5 pages; present results from analysis. Also include descriptive statistics and model selection, if applicable. Reference tables and figures here. Full interpretation of results including pt estimates/CI/pvals/interpretation.

## Discussion

0.5-1pg; interpret results in context of scientific Qs; discuss limitations

## Appendix

Directory: [Github](https://github.com/eweible/BIOS6624.git)

File name: P0/Code/P0_Report.Rmd

Tables and Figures:

```{r, echo=FALSE, fig.cap="(Research Question 1) Agreement between reported booklet time since wake and reported cap time since wake."}
rq1_plot
```

```{r, echo=FALSE, fig.cap="(Research Question 3) Predicted trend of cortisol over time since wake, reported from booklet."}
rq3_plot_c
```

```{r, echo=FALSE, fig.cap="(Research Question 3) Predicted trend of DHEA over time since wake, reported from booklet."}
rq3_plot_d
```

