---
output:
  bookdown::pdf_document2:
    toc: false
    number_sections: false
documentclass: article
geometry: margin=1in
fontsize: 12pt
linestretch: 2
fig.cap: true
header-includes:
  - \usepackage{float}
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
library(ggplot2)
library(reshape2)
library(dplyr)
library(tidyr)
library(powertools)
library(gt)
library(scales)
library(stringr)
dat_prelim <- read.csv(here::here("P2/DataRaw", "PrelimData.csv"))
```

<!-- Title here: -->
# \textit{Alzheimer's Disease and Aging: Aims 1 and 2}

<!-- Name here: -->
Ezra Weible[^1]

[^1]: Biostatistics Masters Student, Colorado School of Public Health, CU Anschutz Medical Campus

```{r correlations, echo=FALSE}
cor_matrix <- cor(dat_prelim, use = "pairwise.complete.obs")
  # Correlations

fig1 <- ggplot(melt(cor_matrix), aes(Var1, Var2, fill = value)) +
  geom_tile() +
  scale_fill_gradient2(
    limits = c(-1, 1)
  ) +
  theme_bw() +
  theme(
    axis.text.x = element_text(angle = 45, hjust = 1),
    axis.text.y = element_text(angle = 0),
    axis.title.x = element_blank(),
    axis.title.y = element_blank()
  ) + 
  labs(title = "Correlation heatmap",
       x = "", y = "") # Correlation visualization
```
```{r aim1_power, echo=FALSE}
r_vals <- seq(0.2, 0.5, by = 0.05)
  # Correlation range
N <- 175
  # Sample size
alpha_markers <- 0.05 / 4   
alpha_outcomes <- 0.05 / 6  
alpha_all <- 0.05 / 24
  # Alpha levels
aim1_results <- data.frame(
  r = r_vals,
  power_alpha_markers = sapply(r_vals, function(r)
    corr.1samp(N = N, rho0 = 0, rhoA = r, alpha = alpha_markers, sides = 2)),
  power_alpha_outcomes = sapply(r_vals, function(r)
    corr.1samp(N = N, rho0 = 0, rhoA = r, alpha = alpha_outcomes, sides = 2)),
  power_alpha_all = sapply(r_vals, function(r)
    corr.1samp(N = N, rho0 = 0, rhoA = r, alpha = alpha_all, sides = 2))
) # Power calculation

aim1_table <- aim1_results %>% 
  mutate(
    power_alpha_markers = round(power_alpha_markers, 3),
    power_alpha_outcomes = round(power_alpha_outcomes, 3),
    power_alpha_all = round(power_alpha_all, 3)
  ) %>% # Round digits
  gt() %>% # Convert to gt
  tab_header(
    title = "Power to Detect Correlations Across \u03B1 Levels",
    subtitle = "N = 175; Two-sided tests; Adjusted \u03B1 for markers, outcomes, and all tests"
  ) %>% # Labels
  cols_label(
    r = "Correlation (r)",
    power_alpha_markers = "Power (\u03B1 = 0.05/4)",
    power_alpha_outcomes = "Power (\u03B1 = 0.05/6)",
    power_alpha_all = "Power (\u03B1 = 0.05/24)"
  ) %>% 
  fmt_percent(
    columns = starts_with("power"),
    decimals = 1
  ) %>% 
  data_color(
    columns = starts_with("power"),
    fn = scales::col_numeric(
      palette = c("#f7fbff", "#6baed6", "#08306b"),
      domain = c(0, 1)
    )
  ) %>% 
  tab_options(
    table.font.size = px(14),
    data_row.padding = px(6),
    heading.align = "center"
  )
```
```{r aim2_power, echo=FALSE}
N <- 175
r_high_vals <- seq(0.2, 0.7, by = 0.1)   
  # Plausible high-amyloid correlations
alpha_markers <- 0.05 / 4
alpha_outcomes <- 0.05 / 6
alpha_all <- 0.05 / 24
  # Alpha levels
rho_diffs <- seq(0.1, 0.5, by = 0.1)
  # Small to large interaction effects
ratios <- list(
  "30_70" = c(0.30, 0.70),
  "40_60" = c(0.40, 0.60),
  "50_50" = c(0.50, 0.50),
  "60_40" = c(0.60, 0.40)
) # Amyloid+:amyloid- ratios

compute_power <- function(N1, N2, rho_low, rho_high, alpha) {
  corr.2samp(n1 = N1, n.ratio = N2 / N1, rho1 = rho_low, rho2 = rho_high, alpha = alpha)
} # Power calculation

results <- list()
for (ratio_name in names(ratios)) {
  props <- ratios[[ratio_name]]
  N_high <- round(N * props[1])   
    # Amyloid+
  N_low  <- N - N_high            
    # Amyloid−
  tmp_list <- list()
  for (rho_high in r_high_vals) {
    for (d in rho_diffs) {
      rho_low <- rho_high - d
      if (rho_low < 0 | rho_low > 0.99) next
        # Skip invalid correlations
      tmp_list[[length(tmp_list) + 1]] <- data.frame(
        ratio = ratio_name,
        rho_high = rho_high,
        rho_low = rho_low,
        diff = d,
        power_alpha_markers = compute_power(N_low, N_high, rho_low, rho_high, alpha_markers),
        power_alpha_outcomes = compute_power(N_low, N_high, rho_low, rho_high, alpha_outcomes),
        power_alpha_all = compute_power(N_low, N_high, rho_low, rho_high, alpha_all)
      )}}
  results[[ratio_name]] <- do.call(rbind, tmp_list)
} # Loop over ratios * r_high * delta_rho * alpha
```
```{r aim2_visualizations, echo=FALSE}
# Combine all results into one data frame
plot_df <- bind_rows(
  lapply(names(results), function(name) {
    results[[name]] %>% mutate(ratio = name)
  })
)

# Convert to long format
plot_long <- plot_df %>%
  pivot_longer(
    cols = starts_with("power_"),
    names_to = "alpha_type",
    values_to = "power"
  ) %>%
  mutate(
    alpha_type = recode(alpha_type,
      "power_alpha_markers"  = "0.05/4",
      "power_alpha_outcomes" = "0.05/6",
      "power_alpha_all"      = "0.05/24"
    ),
    power = power * 100 
  )

fig2 <- ggplot(plot_long, aes(x = rho_high, y = power, color = alpha_type)) +
  geom_line(size = 1.1) +
  geom_hline(yintercept = 75, linetype = "dashed") +
  facet_grid(diff ~ ratio) +
  labs(
    title = "Power vs Amyloid+ Correlation",
    x = "Amyloid+ Correlation (ρ_high)",
    y = "Power (%)",
    color = "Alpha Level"
  ) +
  scale_color_manual(values = c("red", "blue", "darkgreen")) +
  theme_minimal(base_size = 14) +
  theme(
    strip.text = element_text(size = 14, face = "bold"),
    legend.position = "bottom"
  )

summary_80 <- plot_long %>%
  group_by(ratio, diff, alpha_type) %>%
  # keep only scenarios with at least 80% power
  filter(power >= 80) %>%
  # among those, find the smallest rho_high that achieves it
  slice_min(rho_high, with_ties = FALSE) %>%
  ungroup() %>%
  arrange(ratio, diff, alpha_type)

summary_80_wide <- summary_80 %>%
  select(ratio, diff, alpha_type, rho_high) %>%
  pivot_wider(
    names_from  = alpha_type,
    values_from = rho_high,
    names_prefix = "rho_high_at_80pct_"
  ) %>%
  arrange(ratio, diff)

aim2_table <- summary_80 %>% 
  mutate(
    alpha_type = paste0("α = ", alpha_type),
    rho_high   = round(rho_high, 2),
    diff       = round(diff, 2),
    power      = round(power, 1),
    ratio_label = case_when(
      ratio == "30_70" ~ "30/70 Split (Low Prevalence)",
      ratio == "40_60" ~ "40/60 Split (Moderate Imbalance)",
      ratio == "50_50" ~ "50/50 Split (Balanced Groups)",
      ratio == "60_40" ~ "60/40 Split (Moderate Imbalance)",
      TRUE ~ ratio
    )
  ) %>% 
  select(-rho_low) %>%
  arrange(ratio_label, diff, alpha_type) %>% 
  gt(groupname_col = "ratio_label") %>% 
  tab_header(
    title = "Minimum Amyloid+ Correlation (ρ High) Needed for 80% Power",
    subtitle = "Grouped by prevalence ratio, interaction size (Δρ), and α level"
  ) %>% 
  cols_label(
    diff       = html("<div style='white-space:normal;'>Interaction Size (Δρ)</div>"),
    alpha_type = html("<div style='white-space:normal;'>α Level</div>"),
    rho_high   = html("<div style='white-space:normal;'>Minimum ρ High<br>for 80% Power</div>"),
    power      = html("<div style='white-space:normal;'>Power (%)</div>")
  ) %>% 
  fmt_number(
    columns = c(diff, rho_high, power),
    decimals = 2
  ) %>% 
  # data_color(
  #   columns = rho_high,
  #   fn = scales::col_numeric(
  #     palette = c("#08306b", "#6baed6", "#f7fbff"),
  #     domain = range(summary_80$rho_high, na.rm = TRUE)
  #   )
  # ) %>% 
  tab_options(
    table.font.size = px(14),
    data_row.padding = px(6),
    heading.align = "center",
    column_labels.background.color = "#f0f0f0"
  )

aim2_table_wide <- summary_80_wide %>% 
  gt(groupname_col = "ratio") %>% 
  tab_header(
    title = "Smallest \u03C1 High Achieving ≥ 80% Power",
    subtitle = "Across \u03B1 levels and effect size differences"
  ) %>% 
  cols_label(
    diff = "Difference (diff)",
    `rho_high_at_80pct_0.05/4` = "\u03C1 High at 80% (\u03B1 = 0.05/4)",
    `rho_high_at_80pct_0.05/6` = "\u03C1 High at 80% (\u03B1 = 0.05/6)",
    `rho_high_at_80pct_0.05/24` = "\u03C1 High at 80% (\u03B1 = 0.05/24)"
  ) %>% 
  fmt_number(
    columns = starts_with("rho_high"),
    decimals = 2
  ) %>% 
  data_color(
    columns = starts_with("rho_high"),
    fn = scales::col_numeric(
      palette = c("#08306b", "#6baed6", "#f7fbff"),
      domain = range(summary_80_wide %>% select(starts_with("rho_high")), na.rm = TRUE),
      na.color = "white"
    )
  ) %>% 
  tab_options(
    table.font.size = px(14),
    data_row.padding = px(6),
    heading.align = "center"
  )
```

## Data Analysis

Aim 1 will evaluate whether peripheral inflammatory markers are associated with a 1 year decline in memory and cortical thickness in aMCI. The outcomes will be AD-signature cortical thickness (continuous) and five episodic memory outcomes (continuous; (1) 20' Delay, Total Correct, (2) Recognition, D-Prime, (3) 15' Delay, Total Correct, (4) Proportion Correct, (5) 1-Week Delay, Total Correct). All outcomes will be modeled as a 1 year change (Year 1 - Baseline), and models will include baseline value adjustments to allowing coefficients to be interpreted as the effect on the 1 year change. Hypothesis 1a will examine baseline cytokine (IL-6, TNF-$\alpha$) and chemokine (MCP-1, Eotaxin-1) levels as primary predictors; hypothesis 1b will test changes in these markers over the 1 year period. Each inflammatory marker will be analyzed in separate models to avoid multicollinearity, with exploratory joint models. The general model form for aim 1 will be $Outcome_{Y1-Y0} = \beta_0 + \beta_1 Marker_{Y1 \text{ or } Y1-Y0} + \beta_2 Outcome_{Y0} + \beta_n Covariates + \epsilon$ with an alternative hypothesis that $\beta_1 \neq 0$. 

Aim 2 will examine whether amyloid deposition modifies the relationship between inflammation and clinical progression. The primary hypothesis will test whether baseline amyloid deposition interacts with baseline inflammatory markers to predict a one year decline in episodic memory and cortical thickness. The secondary hypothesis will test whether baseline amyloid deposition interacts with the 1 year change in inflammatory markers to predict the one year decline in the outcomes. Models will follow a similar approach to aim 1, but will include baseline amyloid and an interaction. The general model form for aim 2 will be $Outcome_{Y1-Y0} = \beta_0 + \beta_1 Marker_{Y1 \text{ or } Y1-Y0} + \beta_2 Amyloid_{Y0} + \beta_3 Marker_{Y1 \text{ or } Y1-Y0} * Amyloid_{Y0} + \beta_4 Outcome_{Y0} + \beta_n Covariates + \epsilon$ with an alternative hypothesis that $\beta_3 \neq 0$. 

All models will adjust for age, sex, APOE status, and baseline outcome *a priori*. Additional potential covariates (including education, baseline CDR, GDS,BMI, hypercholesterolemia, blood pressure, NSAID use, immune-related health conditions, and cardiovascular risk) will be evaluated as potential confounders or precision variables. This evaluation will rely on simple exploratory checks of association among markers, outcomes, and candidate covariates. Variables that primarily improve precision will be added to the final models, improving power. Interpretation will be cautious in the presence of correlated variables. Given correlations among outcomes and markers, a mild multiple-comparison approach will be implemented. Once empirical correlations are known, more formal adjustments will be explored. For primary analyses, type I error will be controlled with an adjusted Bonferroni approach by dividing $\alpha$ by 4 (markers), 6 (outcomes), and 24 (markers*outcomes), applied separately within each hypothesis.

## Power

Because all longitudinal analyses model 1 year change outcomes adjusted for baseline values, the effect of interest can be expressed as the partial $R^2$ attributable to the predictor of interest. In practice, this can be approximated using the correlation between each inflammatory marker and each outcome. Preliminary data for two change outcomes (change in cortical thickness and change in CLVT) and two baseline inflammatory markers (IL-6 and MCP-1), produced correlations ranging from 0.2-0.7 (*Figure 1*), which informed the range of effect sizes examined. For both aims, power was calculated using the expected sample size of 175 and the three adjusted $\alpha$ levels described above ($0.05/4$, $0.05/6$, $0.05/24$). These analyses provide a conservative estimate of power, as precision variables will reduce unexplained variance in the final models. 

Power for Aim 1 is summarized in *Table 1*. Even for the most conservative $\alpha$ tested, power is above 80\% when the correlation is at least 0.3. Power remains above 80\% for a correlation of 0.25 for $\alpha = 0.05/4$, but power is low for correlations of 0.2 and below across all $\alpha$s. 

As Aim 2 is focused on a continuous x continuous interaction, we simplified to a more conservative approach of treating amyloid deposition as dichotomous: positive or negative. Power calculations varied the correlation in the amyloid-positive group ($\rho_{high}$ = 0.2-0.7), the difference between amyloid groups ($\Delta \rho$ = 0.1-0.5), and the amyloid prevalence ratio (30/70, 40/60, 50/50, 60/40). These scenarios reflect a range of biolgically plausibe interaction effects and group distributions. Power for Aim 2 is summarized in *Table 2* and visualized in *Figure 2*. 

\newpage
## Appendix 

Code Directory: https://github.com/eweible/BIOS6624.git

File name: P2/Code/P2_Report.Rmd

#### Tables and Figures

```{r fig1, echo=FALSE}
fig1
```

```{r tab1, echo=FALSE}
aim1_table
```

```{r fig2, echo=FALSE}
fig2
```

```{r tab2, echo=FALSE}
aim2_table
```

# TODO - check figure and table numbering to be consistent with the grant

# TODO - change values tested for aim 2 to be a smaller set (remove 0.1, etc.)