---
title: "P4_Data_Exploration"
output: html_document
date: "2026-05-04"
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
library(hdrm)
```

```{r}
generate_data <- function(n, rho) {
  hdrm::gen_data(
    n = n,
    p = 20,
    p1 = 5,
    beta = c(seq(1/6, 5/6, by = 1/6), rep(0, 15)),
    family = "gaussian",
    signam = "heterogeneous",
    corr = "exchangeable",
    rho = rho
  )
}
# Usage: 
# set.seed(2025)
# for (i in 1:10000) {
#   dat <- generate_data(n = 250, rho = 0.35)
#   # run selection methods...
# }
```

## Reduced simulation shell example:

The purpose of this file is to sketch out a possible simulation framework that can scale well, and (hopefully) keep your code files clean and maintainable.

This will be a series of functions and objects that will nest together, hopefully in a logically modular way.

## General path

I will present a series of nested functions that automate the running of a complete simulation run for one profile of simulation settings. 

1. An outer function that iterates through the required number of runs.

2. An inner function that handles data generation and the running of analyses

3. A result collection/organization function

I am going to present these in reverse order, which is not the best for conceptual understanding, but better for function assembly.

Using base R piping, not tidyverse. There are a few differences, but if you just use it the normal way you will not know the difference. One difference is how a dataset is indicated (if dealing with a function that does not take the data in the first position with tidyverse: data %>% someFunc(x, y, .) data is indicated by a period. but in the base version i believe it uses an underscore. There are some other restrictions on use, but ther are really edge cases.

## Summarize selection results

One benefit of this analysis is that all models will ultimately be run using the `lm()` function. This presents an opportunity to save coding by developing a uniform function to extract the results we need from the analysis. 

Each iteration will have data generated by `hdrm::gen_data()`, a saturated model will be run, a reduction algorithm will follow and we will end up with a final `lm()` fit object.

```{r message = F, warning = F}
library(hdrm)

# temp data and model object for testing the function
betas <- c(1/6, 1/3, 1/2, 2/3, 5/6, rep(0, 15))

dat_obj <- gen_data(n = 250 
               ,p = 20
               ,p1 = 5
               ,beta = betas
               ,family = 'gaussian'
               ,corr = 'exchangeable'
               ,rho = 0.35
               ) 
dat <- data.frame(y=dat_obj$y, dat_obj$X)

sat_model <- lm(y ~ ., data = dat)
summary(sat_model)

pval_redux <- step(sat_model
                  ,direction = 'backward'
                  ,k = qchisq(1-0.05, 1)
                  )
class(summary(pval_redux)$coefficients)


# results extraction function
harvest <- function(mod_obj      # result model object from the reduction
                   ,param        # simulation target betas
                   ,alpha = 0.05 # crit for coverage and error computation
) {
  
  # extract model results
  estimates <- as.data.frame(summary(mod_obj)$coefficients) 
  estimates <- within(estimates, {
    param <- row.names(estimates)
    })
  CIs <- as.data.frame(confint.default(mod_obj, level = 1-alpha))
  names(CIs) <- c('LCL','UCL')
  CIs <- within(CIs, {
    param <- row.names(CIs)
  })
  
  var_names <- sprintf("V%02d", 1:length(param))
  
  # assemble summaries
  res_dat <- data.frame(
    variables = var_names
   ,true_values = param
  ) |>
    within({
      selected <- ifelse(variables %in% row.names(estimates), 1, 0)
      signif   <- ifelse(variables %in% row.names(estimates[estimates[,4] < alpha,]), 1, 0)
      true_non_zero <- ifelse(variables %in% sprintf("V%02d", 1:5), 1, 0)
    }) |>
    merge(CIs
          ,by.x = 'variables'
          ,by.y = 'param'
          ,all.x = T) |>
    within({
      covered <- ifelse(LCL <= true_values & true_values <= UCL, 1, 0)
    })
    
  # fix non-selected for coverage
  res_dat[is.na(res_dat$covered) == T & 
          res_dat$selected == 0 & 
          res_dat$true_non_zero == 1,'covered'] <- 0 
  res_dat[is.na(res_dat$covered) == T & 
          res_dat$selected == 0 & 
          res_dat$true_non_zero == 0,'covered'] <- 1 
  
  return(res_dat)
}

testit <- harvest(pval_redux, param = betas, alpha = 0.01)
```


## Simu function: generating data and running analyses

In this function I propose to encode the process of running one simulation iteration collecting and returning the results.

```{r}

simfunc <- function(n, rho) {
  
  # create simulated data
  param <- c(1/6, 1/3, 1/2, 2/3, 5/6, rep(0, 15))
  dat_obj <- gen_data(n = n 
                 ,p = 20
                 ,p1 = 5
                 ,beta = param
                 ,family = 'gaussian'
                 ,corr = 'exchangeable'
                 ,rho = rho
                 ) 
  dat <- data.frame(y=dat_obj$y, dat_obj$X)

  # reduction by p-values, AIC, BIC
  sat_model <- lm(y ~ ., data = dat)

  ## p-values
  pval_redux <- step(sat_model
                    ,direction = 'backward'
                    ,trace = 0
                    ,k = qchisq(1-0.05, 1)
                    )
  pval_res <- harvest(mod_obj = pval_redux  # result model object from the reduction
                     ,param   = param       # simulation target betas
                     ,alpha   = 0.05        # crit for coverage and error computation
  )
  pval_res$method <- 'PVAL'
  pval_res$n      <- n
  pval_res$rho    <- rho
  
  ## AIC
  AIC_redux <- step(sat_model
                    ,direction = 'backward'
                    ,trace = 0
                    ,k = 2
                    )
  AIC_res <- harvest(mod_obj = AIC_redux  # result model object from the reduction
                     ,param   = param       # simulation target betas
                     ,alpha   = 0.05        # crit for coverage and error computation
  )
  AIC_res$method <- 'AIC'
  AIC_res$n      <- n
  AIC_res$rho    <- rho

  return(rbind(pval_res, AIC_res))
}

test_sim <- simfunc(n = 250, rho = 0.5)
```




## Dealing with many simulation cases

In our simulation we have two main attributes that are being manipulated: sample size and predictor correlation. One way to handle that would be nesting for-loops, but this can get messy fast and it is easy to mix up your indices. One way to cut down the complexity is a profile matrix:

```{r}

profiles <- expand.grid(
  rho = c(0, 0.35, 0.7)
 ,N   = c(250, 500)
)


profiles

```

With this object we only need to cycle through one loop to find all settings for the current run:

```{r}

# the ith sim:
i <- 3

# settings:
profiles[i,]


```

Now we need a function to iterate through a complete simulation profile.

```{r}

spin <- function(nsim = 500, profile){
  
  stopifnot(nrow(profile) == 1)
  
  res <- list()
  for(iter in 1:nsim) {
    tmp <- simfunc(n = profile[,'N'], rho = profile[,'rho'])
    tmp$iter <- iter
    res[[iter]] <- tmp
  }
  res <- do.call('rbind', res)
  
  return(res)
}
  
test_spin <- spin(nsim = 5, profile = profiles[3,])
```

## The Full Simulation

Defaulting to 10,000 simulations may extend the simulation time, however, estimating a sample size using some reasonable targets may result in a more feasible load (Morris, 2019 - this was uploaded to the week 12 module). For this I will choose to power on the coverage. For this I will use an expected coverage probability of 50% (the worst case in terms of variance). For the SE, I would like to get fairly close, so will use 0.01. This will give reasonable precision in cases of very poor coverage, and much better precision when coverage is closer to 95%. Of course, in critical simulations where precision is critical, higher numbers may be needed and one must give thought to this.

```{r}

# expected coverage
Ep <- 0.5

# required standard error
SEr <- 0.01
nsim <- ceiling(Ep*(1-Ep)/SEr^2)
nsim

# for comparison at 95% coverage and and SE of 0.005 (so we will have better precision than this, where it matters!)
0.95*0.05/(0.005^2)
```

This run took about 4.5 minutes

```{r eval = F}
system.time({
test_spin <- spin(nsim = nsim, profile = profiles[3,])
})
```

The full simulation, as coded (i.e. not the full simulation for the project!) will take about 30 minutes:

```{r eval = F}
res <- list()
for(i in 1:6) {
  profile <- profiles[i,]
  print(profile)
  
  res[[i]] <- spin(nsim = 5, profile = profile)
}

```

If you are in a hurry, the `future` package can make parallel processing fairly easy:

```{r message = F, warning = F}
library(future)
library(future.apply)

n_profiles <- nrow(profiles)

plan(multisession, workers = n_profiles)

set.seed(12345)

system.time({
simres <- future_lapply (1:n_profiles, function(i){
  spin(nsim = nsim, profile = profiles[i,])
}
,future.seed=TRUE)
})
simres <- do.call('rbind', simres)
# stop the cluster
plan(sequential)
```

How many 'workers' can you pick? up to the number of cores on your computer but you should leave at least one free for the operating system to continue to function well. You can get a count of available cores:

```{r}
parallel::detectCores()

```

Compute Coverage probability, by method and simulation setting:

```{r message=F, warning=F}
library(tidyverse)
p_cover <- simres |>
  filter(true_non_zero == 1) |>
  group_by(n, rho, method, variables) |>
  summarise(P_cover = mean(covered)*100)

knitr::kable(
p_cover |> pivot_wider(
  names_from = variables
 ,values_from = P_cover
),
digits = 1)
```



# References

1. Morris TP, White IR, Crowther MJ. Using simulation studies to evaluate statistical methods. Stat Med. 2019;38(11):2074-2102. doi:10.1002/sim.8086


